---
title: "Using socio-demographic data in practice"
author: "Marga Torre"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
    css: styles.css
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T, message=FALSE, warning=FALSE, knitr.purl.inline = TRUE)
```

```{r include=FALSE}
library(haven)
library(tidyverse)
library(knitr)
library(gmodels)
library(ggplot2)
library(openxlsx)
library(skimr)
library(ggplot2)
library(janitor)
library(readr)
library(dplyr)
library(labelled)
library(essurvey)
```

## European Social Survey

The [European Social Survey (ESS)](http://www.europeansocialsurvey.org/about-ess) is an academically driven cross-national survey that has been conducted across Europe since its establishment in 2001. Every two years, face-to-face interviews are conducted with newly selected, cross-sectional samples. You can access the questionnaire from [here](https://www.europeansocialsurvey.org/methodology/ess-methodology/source-questionnaire).

## Read data

Let’s begin by loading the data for Spain, rounds 1–10.

```{r}
data_full <- read_csv("data.csv")
dim(data_full)

```

Alternatively, you can use the [`essurvey` package](https://cran.r-project.org/web/packages/essurvey/index.html) to retrieve the data from the website.

### Select variables

ESS round + ID + socio-demographic profile, including: gender, age of the respondent, years of education, level of education, occupation, household's total net income and citizenship. Please be aware that **some variable names may change from one round to another**.

```{r}
data<- data_full |> select (essround, idno, agea, eduyrs, eisced, gndr, iscoco, isco08, hinctnt, hinctnta)
```

## Data cleaning and data wrangling

### Gender

Generate a new variable named `men` and assign it the following values. Label the categories of the variable and check the results using a table.

-   1=man
-   0=woman
-   other=NA

```{r}
data <- data |>
  mutate(
    male = case_when(
      gndr == 1 ~ 1,  
      gndr == 2 ~ 0,  
      gndr == 9 ~ NA_real_,  
      TRUE ~ NA_real_  
    ),
    male = factor(male, 
                  levels = c(0, 1), 
                  labels = c("female", "male"))
  )

table(data$essround, data$male)
```

### Age

To prepare the continuous age variable, we first need to check the minimum and maximum ages as recorded in the European Social Survey (ESS). The dataset metadata provides this information.

```{r}
data <- data |>
  mutate(
    age = replace(agea, agea > 103, NA)
  )

summary(data$age) 
```

Using the continuous variable `age`, we're going to generate a new variable named `age_rec` with the following categories (levels).

1.  "\<20"
2.  "20-29"
3.  "30-44"
4.  "45-64"
5.  "\>=65"

```{r}
data <- data |>
  mutate(
    age_rec = cut(agea, breaks = c(0, 19, 29, 44, 64, 103),
                  labels = c("<20", "20-29", "30-44", "45-64", ">=65"))
  )

with(data, table(essround, age_rec))
```

### Education

We will now create a continuous variable called `years_schooling` using the variable `eduyrs`.

```{r}
data <- data |>
  mutate(years_schooling = replace(eduyrs, eduyrs > 60, NA))

summary(data$years_schooling)
```

Based on the `_EISCED_` variable, we generate a new variable named `_educ_level_` with the categories listed below.

1.  Low education
2.  Medium education
3.  High education

```{r}
data <- data |>
  mutate(
    educ_level = case_when(
      eisced %in% c(1, 2) ~ "Low education",
      eisced %in% c(3, 4) ~ "Secondary education",
      eisced %in% c(5, 6, 7) ~ "High education",
      eisced %in% c(0, 55, 77, 88, 99) ~ NA_character_,
      TRUE ~ NA_character_
    ) |> factor()
  )


with(data, table(essround, educ_level))
```

### Household's total net income

Now, we generate a scale variable named `hincome` representing income deciles. Note that the variable name changes from Wave 3 to Wave 4.

First, we harmonize the name of the variable

```{r}
data <- data |>
  mutate(hincome = case_when(
    essround <= 3 ~ hinctnt,
    essround >= 4 ~ hinctnta,
    TRUE ~ NA_real_
  )) 
```

As we know, income often has high non-response. Let’s check the proportion of non-valid cases.

```{r}
with(data, table(essround, is.na(hincome)))
```

As we can see, the proportion of non-response is quite high in several rounds. It reaches particularly high levels in Round 1 (around 40%), Round 3 (around 42%), and Round 4 (close to 37%), with substantial levels also observed in later rounds.

Next, we remove non-valid answers from `hincome`

```{r}
data <- data |>
  mutate(hincome = ifelse(hincome > 10, NA, hincome))

with(data, table(essround, hincome))
```

However, **this does not solve the underlying problem**. By excluding these observations, we are removing a substantial portion of the sample, which may affect the representativeness of the data. Therefore, whenever possible, it is advisable to impute these missing values rather than simply deleting them, in order to preserve information and reduce potential bias.

### Occupational categories

Using the variables *iscoco* (waves \<=5) and *isco08* (waves \>5), generate a new variable named `occup` with the following levels:

1.  Managers
2.  Professionals
3.  Technicians
4.  Clerical
5.  Service
6.  Agricultural
7.  Blue-collar

As in the previous case, we first harmonize the name of the variable:

```{r}
data <- data |>
  mutate(occup = case_when(
    essround <= 5 ~ iscoco,
    essround >= 6 ~ isco08,
    TRUE ~ NA_real_
  )) 
```

Next, we remove non-valid entries, define the levels, and label them.

```{r}
data <- data |> 
  mutate(
    occup = replace(occup, occup > 10000, NA),
    occup = cut(occup,
      breaks = c(0, 1999, 2999, 3999, 4999, 5999, 6999, 9999),
      labels = c("managers", "professionals", "technicians", 
                 "clerical", "service", "agricultural", "blue-collar")
    ))
```

Let's check the new variable by round:

```{r}
with(data, table(essround, occup))
```

Finally, we can extract R code from the R Markdown document using the `purl()` function from the `knitr` package. This function takes an R Markdown file as input and generates a pure R script containing only the code chunks from the document. Here's how you can do it:

```{r}
options(knitr.duplicate.label = 'allow')
purl("ESS_es_sociodemographics.Rmd", output = "ESS_es_sociodemographics.R")
```

By extracting the code into a standalone R script, you can easily reuse and source the code in other projects, R scripts, or R Markdown documents. This is especially useful for wrangling tasks, where you might develop a series of data cleaning, transformation, and analysis steps that are applicable across multiple datasets or projects.
