---
title: "2 - Sentiment analysis_SOLVED"
course: Text Mining - MUCSS 25/26
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
instructor: Carmen Torrijos
based_on: tidytextmining.com
editor_options: 
  chunk_output_type: inline
---

# Sentiment analysis

Let's do some sentiment analysis using lexicons/dictionaries. This means we work in the word-level.

## 1. Know our lexicons

Before we start analyzing, let's have a look at some sentiment lexicons. The `tidytext` package provides access to three general-purpose sentiment lexicons:

-   AFINN from Finn Årup Nielsen,

-   Bing from Bing Liu and collaborators, and

-   NRC from Saif Mohammad and Peter Turney.

```{r}
#we install the necessary packages to work with these lexicons
# install.packages("tidytext")
# install.packages("textdata")
```

The three packages have different approaches to sentiment analysis.

```{r}
library(tidytext)

get_sentiments("afinn")
```

Afinn is a regression approach. It gives points to words from 5 to -5.

```{r}
get_sentiments("bing")
```

Bing is a multiclass classification approach on just 2 categories: positive and negative.

```{r}
get_sentiments("nrc")
```

NRC is a multilabel classification approach in two different levels: it combines sentiment and emotions.

## 2. Extracting sentiment from Jane Austen novels

We will be working on a dataset we created in notebook 1: tokenized Jane Austen novels with columns for the books and chapters the word is coming from. Let's create it again:

```{r}
library(janeaustenr)
library(dplyr)
library(stringr)

tidy_austen_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, 
                                regex("^chapter [\\divxlc]", 
                                      ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)

tidy_austen_books
```

### Joy words in Emma

Now that we have our books tokenized by word, let's make a question: What are the most common **joy words** Jane Austen used? Let's take one of the novels, for example, *Emma*.

```{r}
#we set nrc lexicon to a variable, filtering by joy
nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy")

emma_joy_words <- tidy_austen_books %>%
    #we choose the book Emma
    filter(book == "Emma") %>%
    #we combine both lists (all matching rows), NRC and Emma's words
    inner_join(nrc_joy) %>%
    #we count the mentions of each word to find the most frequent
    count(word, sort = TRUE)

emma_joy_words
```

We've got a tibble with words and frequencies from *Emma*, filtered only for joy words.

**Exercise**: We want to make a plot combining the last piece of code with ggplot2. We will follow these steps:

1.  Start from tidy_austen_books

2.  Filter by book *Emma*

3.  Combine with nrc_joy

4.  Count word frequencies and sort them

5.  Filter them by frequency (only mentioned more than X times)

6.  Reorder column word by number of mentions (most frequents on top)

7.  Create the plot with x=n, y=word

Write the three missing lines of code:

```{r}
library(ggplot2)

tidy_austen_books %>%
  filter(book == "Emma") %>%
    #we combine both lists, coincident items in NRC and Emma
    inner_join(nrc_joy) %>%
    #we count the mentions to see the most frequent words
    count(word, sort = TRUE) %>%
    #only words mentioned over 50 times in the novel
    filter(n > 50) %>%
    #we reorder words by number of mentions
    mutate(word = reorder(word, n)) %>%
    #we create the plot with the word (x) and the number of mentions (y)
    ggplot(aes(x = n, y = word)) +
    geom_col() +
    labs(y = NULL)
```

### Sentiment distribution in novels

We can also examine how sentiment changes throughout each novel. First, we divide text in chunks and find a sentiment score for each one of them using the Bing lexicon and [`inner_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html).

```{r}
library(tidyr)

jane_austen_sentiment <- tidy_austen_books %>%
  #find the sentiment for each word using bing
  inner_join(get_sentiments("bing")) %>%
  #divide each book in chunks of 80 lines assigning an index
  count(book, index = linenumber %/% 80, sentiment) %>%
  #we write positive and negative in different columns
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  #we substract positive minus negative to find a net sentiment for each chunk
  mutate(sentiment = positive - negative)

jane_austen_sentiment
```

We've got everything we need to create plots of sentiment by novel. We can plot these sentiment scores across the plot trajectory of each novel following the chunks of 80 lines.

```{r}
library(ggplot2)
#create the plot with x = index (chunks) and y = net sentiment
ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")
```

Remember we are using the sentiment lexicon "Bing" here. Try to imagine the plot of each novel just by looking at the distribution of sentiment.

**Exercise:** Read the plot, locate the points of conflict for each book and think if they correspond to common narrative structures.

## 3. Comparing lexicons in Pride and Prejudice

Let's use the three lexicons, AFINN, Bing and NRC, applied to the same novel, to compare the narrative distribution of sentiment in the three of them.

First of all, let's filter the book we are interested in from the general data frame of novels.

```{r}
pride_prejudice <- tidy_austen_books %>% 
  filter(book == "Pride & Prejudice")

pride_prejudice
```

Now, let's extract the net sentiment with each lexicon for chunks of 80 lines.

```{r}
#for AFINN is different. We need to summarise quantities to get net sentiment.
afinn <- pride_prejudice %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(index = linenumber %/% 80) %>%
  #we summarise the sentiment of each chunk
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "AFINN")

afinn
```

```{r}
#for Bing and NRC we can do it in one step using bind_rows to paste them together
bing_and_nrc <- bind_rows(
  #Bing
  pride_prejudice %>% 
    #we get sentiments from bing
    inner_join(get_sentiments("bing")) %>%
    #we create the column for bing
    mutate(method = "Bing et al."),
  #NRC
  pride_prejudice %>% 
    #we get sentiment from nrc 
    inner_join(get_sentiments("nrc") %>% 
                 #we filter just sentiment, not emotions
                 filter(sentiment %in% c("positive", 
                                         "negative"))
    ) %>%
    #we create the column for nrc
    mutate(method = "NRC")) %>%
  #we divide in chunks of 80 lines
  count(method, index = linenumber %/% 80, sentiment) %>%
  #we write positive and negative in different columns
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>% 
  #we extract net sentiment by substraction
  mutate(sentiment = positive - negative)

bing_and_nrc
```

We now have an estimate of the net sentiment (positive - negative) in each chunk of the novel text for each sentiment lexicon.

-   Remember that the column "sentiment" is in Afinn a sum of values.

<!-- -->

-   Remember that the column "sentiment" in Bing and NRC is a substraction of positive words - negative words.

And finally, let's bind the three of them together and visualize them in a plot.

```{r}
#bind the three of them
bind_rows(afinn, 
          bing_and_nrc) %>%
  #make the plot with x=index (chunks), y=sentiment and fill by lexicon (method)
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")
```

Now you can see how different narrative plot can be just depending on the sentiment lexicon used. To read the plot, look at:

-   Higher and lower positive values.

-   Larger blocks of similar sentiment.

-   Similarity in points of conflict and narrative arc.

### Lexicons bias and risks

We can not trust the lexicon 100%. Generally, there are three **possible biases** in sentiment lexicons:

-   The lexicon is biased towards negative sentiment (more negative than positive words)

-   The lexicon is biased towards positive sentiment (more positive than negative words)

-   The lexicon is very different in style or context from the text we are analyzing. For instance: a lexicon built on Twitter data used to analyze 19th century literature.

Let's see which case is Bing:

```{r}
get_sentiments("bing") %>% 
  count(sentiment)
```

Let's see which case is NRC:

```{r}
get_sentiments("nrc") %>% 
  filter(sentiment %in% c("positive", "negative")) %>% 
  count(sentiment)
```

As it follows a regression model, the case of Afinn is different, but we can still see how many words it has for each value.

```{r}
get_sentiments("afinn") %>% 
  count(value)
```

Summing up: 1598 negative words and 873 positive words. Only one neutral word.

**Exercise**: which is the neutral word in Afinn?

```{r}
get_sentiments("afinn") %>% 
  filter(value == 0)

```

### Most common positive and negative words

Let's find out how much each word in the pack of novels contributes to each sentiment. This is: how many times each word appears in the novels, and which sentiment it is associated to:

```{r}
#create a data frame with word, sentiment and number of mentions
bing_word_counts <- tidy_austen_books %>%
  #get sentiment from bing
  inner_join(get_sentiments("bing")) %>%
  #count the number of mentions for each word
  count(word, sentiment, sort = TRUE)

bing_word_counts
```

Let's make a plot to see this in perspective.

```{r}
bing_word_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 15) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)
```

Here's how we discover, for example, that the word "miss" is bringing a lot of negative sentiment to the analysis. How can we fix this?

We can add "miss" to an existent stopwords list. Remember that we've got one in `tidytext` built upon 3 different lexicons (SMART; snowball and onix) that you can find directly in `stop_words`.

```{r}
stop_words
```

Now we know the aspect of the tibble, there are two columns to fill: word and lexicon it comes from. For the case of "miss", in the lexicon column we can write just "custom".

```{r}
custom_stop_words <- bind_rows(tibble(word = c("miss"),  
                                      lexicon = c("custom")), 
                               stop_words)

custom_stop_words
```

**Exercise:** create the same dataframe and plot, this time without "miss". Follow these steps:

1.  Filter custom stopwords
2.  Create the data frame again
3.  Create the plot again

```{r}
tidy_austen_books <- tidy_austen_books %>%
  #we use custom_stop_words instead of stop_words
  anti_join(custom_stop_words)
```

```{r}
bing_word_counts <- tidy_austen_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE)

bing_word_counts
```

```{r}
bing_word_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 15) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)
```

### Most negative chapters

Next question could be: what are the most negative chapters in each of Jane Austen's novels? To answer this question we need to extract average sentiment by chapter. The real question is: **which chapter has the highest proportion of negative words?**

We do that in a few steps:

-   First, we filter negative words from Bing.

-   We make a data frame with number of words per chapter.

-   We find the number of negative words in each chapter and divide by the total words in chapter.

```{r}
#filter negative words from Bing
bingnegative <- get_sentiments("bing") %>% 
  filter(sentiment == "negative")

bingnegative
```

```{r}
# make a dataframe (wordcounts) with number of words per chapter
wordcounts <- tidy_austen_books %>%
  group_by(book, chapter) %>%
  summarize(words = n())

wordcounts
```

```{r}
#find the number of negative words by chapter and divide by the total words in chapter
tidy_austen_books %>%
  #semi_join: returns all words in books with a match in bingnegative
  semi_join(bingnegative) %>%
  #group by book and chapter to summarize how many negative words by chapter
  group_by(book, chapter) %>%
  summarize(negativewords = n()) %>%
  #left_join keeps all words in wordcounts and makes a dataframe
  left_join(wordcounts, by = c("book", "chapter")) %>%
  #create a column in the dataframe with the ratio
  mutate(ratio = negativewords/words) %>%
  #we don't want chapters 0 because they're just title and author
  filter(chapter != 0) %>%
  #we select the highest ratios
  slice_max(ratio, n = 1) %>% 
  ungroup()
```

**Exercise:** print one or two of this chapters to see with your eyes if they really are so negative.

```{r}
tidy_austen_books %>%
  filter(book == "Mansfield Park") %>%
  filter (chapter == 45)
```

## 4. Wordclouds ☁️

Wordclouds are a very easy and powerful tool to visualize language and concepts. Here we can see the most common words in Jane Austen books, using directly the function `wordcloud`:

```{r}
# install.packages("wordcloud")
```

**Note**: expect a lot of warnings when painting wordclouds. It's normal, R is tuning the size of the words to make them fit in the cloud.

**Note:** for wordclouds, filtering stopwords is particularly important, as they can really distort the cloud.

```{r}
library(wordcloud)

tidy_austen_books %>%
  #we filter stopwords
  anti_join(custom_stop_words) %>%
  #we count words
  count(word) %>%
  #we use the wordcloud function
  with(wordcloud(word, n, max.words = 60))
```

This is in black and white, so let's put some colour into it. First, we install the package:

```{r}
# install.packages("RColorBrewer")
```

Second, we recreate the word cloud with the new possibilities. There are different ways of introducing colors in word clouds. You can specify colors using either a color name, a hexadecimal color code, or a vector of color values. For example:

-   We write a vector of colors to be combined: `wordcloud(words, freq, colors = c("red", "blue")`

-   Same with hexadecimal color codes: `wordcloud(words, freq, colors = c("#00FF00", "#FF00FF")`

-   We use a number of colors from a pre-defined palette: `wordcloud(words, freq, colors = brewer.pal(8, 'Dark2'))`

Let's try the palette option:

```{r}
library(RColorBrewer)

#set the colors from a brewer palette. 8 is the number of colors used from Dark2 palette.
colors = brewer.pal(10, 'Dark2')

tidy_austen_books %>%
  anti_join(custom_stop_words) %>%
  count(word) %>%
  #we add colors as an argument
  with(wordcloud(word, n, max.words = 90, colors = colors))
```

### Comparing two wordclouds in one ☁️☁️

Let's compare positive wordcloud with negative wordcloud. To be able to plot two wordclouds in one, we need to install reshape2.

```{r}
# install.packages("reshape2")
```

```{r}
library(reshape2)

word_matrix <- tidy_austen_books %>%
  #we get sentiments
  inner_join(get_sentiments("bing")) %>%
  #we count word mentions
  count(word, sentiment, sort = TRUE) %>%
   #we convert to a matrix: rows = words, columns = sentiments, values = counts (0 if missing)
  acast(word ~ sentiment, value.var = "n", fill = 0)

head(word_matrix)
```

```{r}
#we paint two wordclouds in one using two different colors
#the colors are assigned in the same order as the columns in the matrix
comparison.cloud(word_matrix, colors = c("red", "blue"),
                   max.words = 100)
```

**Exercise**: repeat this wordcloud but only restricted to one book from Jane Austen, at your choice.

```{r}
library(reshape2)

tidy_austen_books %>%
  filter(book == "Persuasion") %>%
  #we get sentiments
  inner_join(get_sentiments("bing")) %>%
  #we count word mentions
  count(word, sentiment, sort = TRUE) %>%
  #we create the matrix
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  #we paint two wordclouds in one using two different colors
  comparison.cloud(colors = c("red", "blue"),
                   max.words = 100)
```

## 5. Working sentence-level

So far, we have explored how to analyze sentiment at the word level. This approach can be useful depending on the project, for instance, if you want to create a word cloud.

Now, we are going to look at how to handle sentiment analysis at the **sentence or short text level.**

To do that, we need a library called `sentimentr`.

```{r}
#first, we install the package
# install.packages("sentimentr")
```

```{r}
#we try some sentences to see how it works
library(sentimentr)
texts <- c(
  "I love this book, it is amazing, super cool!",
  "I hate waiting for the train.",
  "The movie was okay, not bad but not great."
)

# calculate sentiment using function sentiment_by
texts_sentiments <- sentiment_by(texts)

texts_sentiments
```

However, this output can be a bit confusing. Let's select only what we are interested in and create a tibble with sentence and sentiment.

```{r}
#we create a tibble with two columns, sentence and sentiment
final_sentiment <- tibble(sentence = texts, sentiment = texts_sentiments$ave_sentiment)


final_sentiment
```

Sentence-level sentiment analysis can look more modern or accurate, but it is not perfect either. Always look for the best option for your specific project.

## Summary

We have learned so far:

-   How to use sentiment lexicons with different approaches

-   How to analyze sentiment in the word-token level

-   How to extract net sentiment of a chunk of lines

-   How to plot the sentiment of a corpus to make a guess at its narrative distribution

-   How to identify words that are contributing the most to sentiment analysis and how to fix mistakes

-   How to find the most negative chapters in a book

-   How to plot word clouds in black and white and color to illustrate our analysis

-   How to analyze sentiment in the sentence or short text level
