---
title: "case_study_mock_task"
author: "Amelia Shuler"
format: html
editor: visual
---

## Setup

Let's perform a small simulation before delivery using the starwars dataset from the {dplyr} package.

```{r}
#| message: false
library(dplyr)
library(tibble)
```

We have different variables of the Star Wars characters, with characteristics of their hair, skin, height, name, etc.

```{r}
starwars
```

## Exercise 1

How many characters are stored in the database? How many characteristics have been measured for each one?

```{r}
nrow(starwars)
ncol(starwars)
```

## Exercise 2

Extract in two different variables names and ages the corresponding variables from the table. What type is the variable name? And the variable birth_year?

```{r}
names <- starwars$name
ages <- starwars$birth_year
class(names)
class(ages)
```

## Exercise 3

Obtain the vector of names of the characters sorted from oldest to youngest.

```{r}
names[order(names, decreasing = TRUE)]
```

## Exercise 4

Get help from the unique() function. Use it to find out what modalities the qualitative variable for eye color has. How many different ones are there?

```{r}
eye_color <- unique(starwars$eye_color)
length(eye_color)
```

## Exercise 5

Are there ANY missing values in the eye color variable?

```{r}
any(is.na(eye_color))
sum(is.na(eye_color)) # if > 0, then yes
```

# Exercise 6

Calculate the mean and standard deviation of the height and weight variables (watch out for missing values).

```{r}
mean(starwars$height, na.rm = TRUE)
sd(starwars$height, na.rm = TRUE)
mean(starwars$mass, na.rm = TRUE)
sd(starwars$mass, na.rm = TRUE)
```

Define a new tibble with these two variables and incorporate a third variable called “BMI” that calculates the body mass index. Incorporate with \$ \$ the formula used for BMI.

```{r}
new_tb <- tibble("height" = starwars$height, "weight" = starwars$mass, "BMI" = weight / (height/100^2))
new_tb
```
