---
title: "full poisson"
author: "Emma Perez"
date: "2025-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# RECAP OF POISSON REGRESSION ASSUMPTIONS

1. Poisson Response: The response variable is a count per unit of time or space, described by a Poisson distribution.
2. Independence: The observations must be independent of one another.
3. Mean=Variance: By definition, the mean of a Poisson random variable must be equal to its variance.
4. Linearity The log of the mean rate, log(λ), must be a linear function of regressors.

# Load data ----

Infidelity data, known as Fair's Affairs. Cross-section data from a survey conducted by Psychology Today in 1969.

```{r}
# install.packages("datasets")
data(Affairs, package = 'AER')
```


- affairs: is the number of extramarital affairs in the past year and is our response variable
- gender
- age
- yearsmarried
- children (yes or no)
- religiousness (1 - anti, 5 - very)
- education
- rating self rating of the marriage 1 (very unhappy) to 5 (very happy)
- and some others

```{r}
summary(Affairs)
```

Have a close look at the distribution of our target variable: 75% of our observations are 0. 

```{r}
Affairs$occupation <- as.factor(Affairs$occupation)
```


# EDA -----

```{r}
library(DataExplorer)

plot_intro(Affairs)
```

```{r}
# histogram
hist(Affairs$affairs)

#density plot 
success <- min(Affairs$affairs):max(Affairs$affairs)

plot(density(Affairs$affairs),ylim=c(0,0.5), xlim = c(0,15))+
lines(success,dpois(success, lambda=mean(Affairs$affairs)),col="green")
```
Check the target variable and its relation to other vars

```{r}
plot_histogram(Affairs)

plot_boxplot(Affairs, by = "affairs")
```

There might be a relationship between rating and having an affair because the box plots scale to the right. Maybe years married and affairs since they tend to have a small relationship.

```{r}
#gender
boxplot(affairs ~ gender, data = Affairs, main = "Number of Affairs by Gender", ylab = "Number of Affairs", xlab = "Gender")
wilcox.test(affairs ~ gender, data = Affairs)
```

We can determine that gender is not statistically significant, but with non-normal distribution these tests are not as reliable.

```{r}
# children
boxplot(affairs ~ children, data = Affairs, main = "Number of Affairs by Children", ylab = "Number of Affairs", xlab = "Children")
wilcox.test(affairs ~ children, data = Affairs)
```

With children, we can see it is statistically significant.

```{r}
library(GGally)
ggcorr(Affairs, label = T, method = c("all.obs","spearman"))
```
Most related vars to the target are: years married and rating of marital satisfaction

but are very weak

Check mean and variance of target

```{r}
mean<- mean(Affairs$affairs)
var <- var(Affairs$affairs)

var/mean
```

variance is 7 times bigger than the expected value of y. We can guess we have a problem of dispersion.

# split data 

```{r}
training.samples <- Affairs$affairs %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- Affairs[training.samples, ]
test.data <- Affairs[-training.samples, ]
```

Simple Cross Validation 80 / 20

# Train and compare models

Let's train a linear and a poisson model to start with

```{r}
formula <- affairs ~ age + yearsmarried + religiousness + rating + rating*religiousness + children 
```

We are going to train a linear model to see how it compares to a Poisson regression.

```{r}
# Train a LM model
lm_model <- lm(formula, data = train.data)

summary(lm_model)

# Train a poisson model
poisson_model <- glm(formula, family = poisson, data = train.data)

poisson_model
```

From the linear model we see that rating and years married are the important variables.

This is the results of the poisson model. The residual deviance is lower than the null deviance, so it is better than a random model. However, it is much higher than the degrees of freedom, which means it does not fit the model.
 
Compare models using Achaiche information criteria (AIC)

```{r}
aic_results <- AIC(lm_model, poisson_model)

# Show results
list(AIC = aic_results)

```
Because our AIC is lower for the poisson model, we can see that the poisson is better than a linear model. 

Lower AIC in Poisson -> better model than linear for this data

Compare MAE for both models

```{r}
lm_preds <- predict(lm_model, type = "response", newdata = test.data)
poisson_preds <- predict(poisson_model, type = "response",  newdata = test.data)

lm_MAE <- mean(abs(test.data$affairs - lm_preds))
poisson_MAE <- mean(abs(test.data$affairs - poisson_preds))

lm_MAE
poisson_MAE
```

This is the residual of errors. Mean absolute error. Comparing them, neither are close to the correct answer of 1.5.

We can also tabulate predictions vs actual values like in a confusion matrix

```{r}
table(true = test.data$affairs, pred = round(poisson_preds))

```
Most people have zero numbers of affairs. But we are predicting a lot of values that should be zero. The poisson regression expects there to be more values than zeros, but in our data this is not the case.

```{r}
# Calculate Pearson residuals
fitted_values <- fitted(poisson_model)
pearson_residuals <- residuals(poisson_model, type = "pearson")

# Plot residuals vs fitted values
plot(fitted_values, pearson_residuals,
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```
We want to have a cloud of dots around the line, rather than patterns.

If residuals show a pattern, additional predictors or transformations may be needed.(Random scatter around zero suggests a well-fitting model)

One possible remedy is to incorporate interaction terms, capturing complex relationships between predictors.

This may improve model fit, but can lead to overfitting.

# Overdispersion -----

```{r}
# Extract residual deviance
residual_deviance <- poisson_model$deviance

# Extract residual degrees of freedom
residual_df <- poisson_model$df.residual

# Compute dispersion factor
dispersion_factor <- residual_deviance / residual_df

dispersion_factor
```

Quite high dispersion factor! The variance is very high in the model, and we have a problem with overdispersion. It should be lower than 1.5.

```{r}
# Overdispersion test
library(AER)
dispersiontest(poisson_model)
```
Ho: There is no overdispersion

Seems there is overdispersion (almost 0 p-value)

The p-value is very low, so we reject the null hypothesis.

# Quasipoisson and negative binomial alternatives ----

Quasi-poisson model

```{r}
# Train a poisson model
qpoisson_model <- glm(formula, family = quasipoisson, data = train.data)

qpoisson_model

# Compute predictions and MAE 
qpois_preds <-predict(qpoisson_model, type = "response", newdata = test.data)
qpois_MAE <- mean(abs(test.data$affairs - qpois_preds))

qpois_MAE

```
Same coefficients as Poisson, but corrected standard errors.
AIC, BIC cannot be used with Quasi-Poisson regression.

Picking a new type of model. Quasi-poisson has a correction for the overdispersion.

The residual deviance has decreased a lot. The fit of the model is getting better, but the predictions will probably not get better because of the nature of the target variable.

Negative - binomial model


```{r}
# Negative binomial model 
library(MASS)
negbin_model<-glm.nb(formula, data = train.data)
negbin_model

# Compute predictions and MAE 
nb_preds <-predict(negbin_model, type = "response", newdata = test.data)
nb_MAE <- mean(abs(test.data$affairs - nb_preds))

nb_MAE
```

The residual deviance here is lowest, meaning the negative binomial is the best fit for this model.

# Zero-inflated models -----

ZI Poisson 

```{r}
library(pscl)
zip <- zeroinfl(formula, data = train.data, dist="poisson")

zip_preds <- predict(zip, type = "response", newdata = test.data)
zip_mae <- mean(abs(test.data$affairs - zip_preds))

zip_mae
```


ZI Negative Binomial 
```{r}
zi_nb <- zeroinfl(formula, data = train.data, dist="negbin")

zi_nb_preds<-predict(zi_nb, type = "response", newdata = test.data)
zinb_mae <- mean(abs(test.data$affairs - zi_nb_preds))

zinb_mae

```


# Compare models ----

In terms of MAE - we dont see much improvement in prediction accuracy

```{r}
list(
  LinearModel = list(MAE = lm_MAE),
  PoissonModel = list(MAE = poisson_MAE),
  QuasiPoisson = list(MAE = qpois_MAE),
  NegBin = list(MAE = nb_MAE),
  ZIP = list(MAE = zip_mae),
  ZI_NB = list(MAE =zinb_mae)
)
```

We can see there are no differences in the predictions. This is a very rare event that although the model is getting better, the predictions are staying the same.

Our target variable is what it is.

In terms of AIC - our models have been improving a lot!

```{r}
list(
  LinearModel = AIC(lm_model),
  PoissonModel = AIC(poisson_model),
 # QuasiPoisson = AIC(), AIC is not valid here
  NegBin = AIC(negbin_model),
  ZIP = AIC(zip),
  ZI_NB = AIC(zi_nb)
)
```
Better statistical fit (AIC) does not necessarily mean better prediction accuracy. 
Zero-inflated models describe the data-generating process better, especially the zeros, but they do not magically predict rare counts more accurately :(

Rare event prediction is inherently hard!!


# Using caret for cross validation and model comparison

Caret does not support zero inflated models, so we will go ahead with the rest. 

```{r}
library(MASS)
library(caret)
library(e1071)


set.seed(123)
#CV setup: cv with 10 folds
train_control <- trainControl(method = "cv", number = 5)

#linear
lm_fit <- train(formula, data = train.data, 
                method = "lm", 
                trControl = train_control)

#poisson
poisson_fit <- train(formula, data = train.data, 
                     method = "glm", family = poisson(),
                     trControl = train_control)

#quasi poisson model (just change the family)
qpois_fit <- train(formula, data = train.data, 
               method = "glm", family=quasipoisson(),
               trControl = train_control)

#negative binomial 
neg_bin_fit <- train(formula, data = train.data, 
               method = "glm.nb",
               trControl = train_control)


```

Now we use resamples() to collect the results from our cv process into one object. 
It just collects results already computed inside each model and put it together.

```{r}
results <- resamples(list(linear = lm_fit, poisson = poisson_fit, qpois = qpois_fit, negbin = neg_bin_fit))
summary(results)

bwplot(results, metric = "MAE")

```
From cv training, it seems that negbin provides the best MAE but differences are very small!


Now we can compute our predictions and metrics on the *test set*, and select our final model. 


