---
title: "C1"
author: "Alejandro Llorente"
date: "26/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
data("marketing", package = "datarium")
```

```{r}
head(marketing)
```

Extending and scaling variables

```{r}
marketing2 <- marketing %>%
  mutate(
    youtube2 = youtube^2,
    youtube3 = youtube^3,
    facebook2 = facebook^2,
    facebook3 = facebook^3,
    newspaper2 = newspaper^2,
    newspaper3 = newspaper^3,
    you_face = youtube*facebook,
    you_news = youtube*newspaper,
    face_news = facebook*newspaper
  )

# scale data
marketing2<-as.data.frame(scale(marketing2))
  
```

Computing linear model with additional data

Defining the model we ran previously

```{r}
modellm<-lm(sales ~ .,data=marketing2)
round(coefficients(modellm),3)
```

```{r}
summary(modellm)
```

We should remove the ones that do not have significant codes. We could do this manually, but we will ask an algorithm to do it for us.

## Glmnet

Now let's use glmnet to apply regularization!

The glmnet package fits regularized regression models (like Lasso, Ridge, and Elastic Net).

Fits a set of models for a sequence of λ values that you specify (our grid)

alpha=1 is the lasso penalty, and alpha=0 the ridge penalty.

### Ridge Regression

```{r}
library(glmnet)

X<-marketing2 |> select(-sales)
Y<-marketing2 |> select(sales)

# grid with lambda values to try
# usually, we want to provide R with a wide range of values
grid = 10^seq(10, -2, length = 100) 
# length = 100 provides 100 values
# 10 and -2 are the powers, so you get really small numbers and really really big numbers

# see grid of numbers
options(scipen = 999)
grid

#ridge regression (alpha=0, search for best lambda)
# lambda = grid takes lambda from the grid defined above
modelR<-glmnet(x=X, y=Y[,1], lambda=grid, alpha=0)
```

The model is "selecting for the best lambda." We will use a hypothetically selected lambda value of 0.05. (Usually it would be more like 0.0590274).

### Check coefficients for a given lambda

```{r}
# Get coefficients at λ = 0.05
# s is indicating the lambda we want to use
# this is just an example, we don't need to run this for real
coef(modelR, s = 0.05)
```

Look for the best lambda value \> model with the maximum dev.ratio (like R2)

```{r}
library(broom)
MR<-tidy(modelR) # makes the output of the model more cleanly
MR |>
  filter(dev.ratio == max(MR$dev.ratio)) |> 
  mutate(across(c(estimate,dev.ratio), round, 3))
# this is showing us that the best lambda is 0.01
```

Estimate is the estimated coefficient for each variable. The variables with smaller coefficients are the ones that are less important for the model.

R is trying all the models for the various lambdas using the whole dataset and tells us the best lambda as results.

## Cross-Validation

Now lets use cross validation to select the best lambda

cv.glmnet() Performs k-fold cross-validation (default: 10-fold) to automatically find the optimal λ.

It internally calls glmnet() multiple times for different λ’s and partitions of the data.

lambda.min → λ giving the minimum cross-validation error.

### Lasso Regression

```{r}
#lasso regression (alpha=1, search for best lambda) and do CV
modelL<-cv.glmnet(x=as.matrix(X), y=Y[,1], lambda=grid, alpha=1)

#extract coefficients
coef(modelL, s = "lambda.min") # telling R to use the lambda that minimizes the error

#compute predictions for 10 observations
predict(modelL, newx=as.matrix(X)[1:10,], s="lambda.min")


```

The variables that are just a dot are the variables R has removed.

NOTE\*\*

In glmnet, the deviance is the model’s measure of lack of fit — like the residual sum of squares

For each value of λ (the regularization parameter), the model has a certain deviance. The dev.ratio is the fraction of deviance explained by the model relative to the null model (a model with only the intercept).
