---
title: "Regularization - Caret example"
author: "Emma Perez"
date: "10/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Load data

```{r}
library(tidyverse)
data("marketing", package = "datarium")
head(marketing)
```

Expand data - feature generation

```{r}
marketing2 <- marketing |>
  mutate(
    youtube2 = youtube^2,
    youtube3 = youtube^3,
    facebook2 = facebook^2,
    facebook3 = facebook^3,
    newspaper2 = newspaper^2,
    newspaper3 = newspaper^3,
    you_face = youtube*facebook,
    you_news = youtube*newspaper,
    face_news = facebook*newspaper
  )

```

Scale data *(ESSENTIAL in shrinkage methods as penalty is applied on coefficients!!)*

```{r}
# separate target and explanatory vars
sales <- marketing2$sales

exp_vars <- marketing2 |> select(-sales)
exp_vars_scaled <-as.data.frame(scale(exp_vars))

# put all vars back together
df <- cbind(sales, exp_vars_scaled)
```

# Data split

This splits the data for cross-validation.

```{r}
library(caret)
training.samples <- marketing2$sales %>% 
  createDataPartition(p = 0.8, list = FALSE) #proportion of the split

train.data  <- df[training.samples, ]
test.data <- df[-training.samples, ]
```

# Train models using Caret

We will use 10-fold cross validation to look for the best values of lambda in Lasso and Ridge, and also for alpha in ElasticNet.

-   The dataset is split into 10 subsets (or folds)
-   The model is trained on 9 folds and tested on the remaining fold.
-   This process repeats 10 times, each time using a different fold as the test set.

## Lambda possible values

```{r}
# This creates a sequence of 100 possible lambda values (a broad range) - those to try during training
lambda <- 10^seq(-3, 3, length = 100)
```

For each lambda value, the model is trained and evaluated on all 10 folds. The lambda value with the lowest average error will be chosen as the best lambda

## Train Ridge regression with caret (alpha = 0)

```{r}
ridge <- train(
  sales ~., # input data
  data = train.data, # specify data
  method = "glmnet", # specify method
  trControl = trainControl("cv", number = 10), # specify how you want the cross-validation to be performed, cv is cross validation, number is number of folds
  tuneGrid = expand.grid(alpha = 0, lambda = lambda) # set of lambda values to test (we provide them), R is performing parameter tuning
)
```

```{r}
# Check best parameters 
# name of model $ save best parameters
ridge$bestTune
```

```{r}
# Get the best model's coefficients
best_lambda <- ridge$bestTune$lambda
ridge_coefficients <- coef(ridge$finalModel, s = best_lambda)

ridge_coefficients

ridge$finalModel$lambda # 100 lambdas tested 
```

## Train Lasso regression with caret (alpha = 1)

```{r}
lasso <- train(
  sales ~., data = train.data, method = "glmnet",
  trControl = trainControl("cv", number = 10),
  tuneGrid = expand.grid(alpha = 1, lambda = lambda)
)

```

Some missing values were found when running the Lasso model. It's fine.

```{r}
# Get the best model's coefficients
lasso_best_lambda <- lasso$bestTune$lambda
lasso_coefficients <- coef(lasso$finalModel, s = best_lambda)

lasso_coefficients
```

We removed too many variables and now only have 2. We would need to refine this.

## Train ElasticNet with caret

```{r}
elastic <- train(
  sales ~., data = train.data, method = "glmnet",
  trControl = trainControl("cv", number = 10),
  tuneLength = 10 #number of combinations to test (we do not provide values, caret chooses)
  )
```

Here, alpha will vary between 0 and 1, allowing Elastic Net to balance Ridge and Lasso. If we don't provide values for alpha, R will be able to try various ones.

tuneLength = 10 tells caret to automatically generate a grid of 10 values for alpha and lambda, which will adjust the penalty mix and strength based on cross-validation.

```{r}
elastic$bestTune
```

```{r}
# Get the best model's coefficients
elastic_best_lambda <- elastic$bestTune$lambda
elastic_coefficients <- coef(elastic$finalModel, s = best_lambda)

elastic_coefficients
```

## Train standard linear regression with caret

```{r}
linear<-train(sales ~., data = train.data,
              method = 'lm',
              metric =  "RMSE",
              trControl = trainControl("cv", number = 10)
              )
```

```{r}
# Get the best model's coefficients
linear_coefficients <- coef(linear$finalModel)

linear_coefficients
```

# Comparing models

```{r}
models <- list(ridge = ridge, 
               lasso = lasso, 
               elastic = elastic, 
               linear=linear)

resamples(models) %>% summary()
```

# Make predictions (test data)

To make predictions with the trained Lasso and Elastic Net models in caret, you can use the predict() function, specifying the new data you want to predict on.

The predict() function automatically uses the best-tuned parameters (best lambda and alpha for each model) stored in the finalModel within the caret object.

```{r}
elastic_predictions <- predict(elastic, newdata = test.data)

ridge_predictions <- predict(ridge, newdata = test.data)

lasso_predictions <- predict(lasso, newdata = test.data)

linear_predictions <- predict(linear, newdata = test.data )
```

# Compute MSE on test data

Comparison of each model's error on the test data, helping you to identify which model performs best

```{r}
mse <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

# Calculate MSE for each model
ridge_mse <- mse(test.data$sales, ridge_predictions)
lasso_mse <- mse(test.data$sales, lasso_predictions)
elastic_mse <- mse(test.data$sales, elastic_predictions)
linear_mse <-  mse(test.data$sales, linear_predictions)

ridge_mse
lasso_mse
elastic_mse
linear_mse
```
