---
title: "Regularized GLMs"
format: html
editor: visual
---

# RECAP OF POISSON REGRESSION ASSUMPTIONS

1.  Poisson Response: The response variable is a count per unit of time or space, described by a Poisson distribution.
2.  Independence: The observations must be independent of one another.
3.  Mean=Variance: By definition, the mean of a Poisson random variable must be equal to its variance.
4.  Linearity The log of the mean rate, log(λ), must be a linear function of regressors.

# Load data ----

Infidelity data, known as Fair's Affairs. Cross-section data from a survey conducted by Psychology Today in 1969.

```{r}
# install.packages("datasets")
data(Affairs, package = 'AER')
library(tidyverse)
library(data.table)
library(tidyverse)
library(caret)
set.seed(1234)
```

-   affairs: is the number of extramarital affairs in the past year and is our response variable
-   gender
-   age
-   yearsmarried
-   children (yes or no)
-   religiousness (1 - anti, 5 - very)
-   education
-   rating self rating of the marriage 1 (very unhappy) to 5 (very happy)
-   and some others

```{r}
summary(Affairs)
```

Have a close look at the distribution of our target variable: 75% of our observations are 0.

```{r}
Affairs$occupation <- as.factor(Affairs$occupation)
```

# EDA -----

```{r}
library(DataExplorer)

plot_intro(Affairs)
```

```{r}
# histogram
hist(Affairs$affairs)

#density plot 
success <- min(Affairs$affairs):max(Affairs$affairs)

plot(density(Affairs$affairs),ylim=c(0,0.5), xlim = c(0,15))+
lines(success,dpois(success, lambda=mean(Affairs$affairs)),col="green")
```

Check the target variable and its relation to other vars

```{r}
plot_histogram(Affairs)

plot_boxplot(Affairs, by = "affairs")
```

There might be a relationship between rating and having an affair because the box plots scale to the right. Maybe years married and affairs since they tend to have a small relationship.

```{r}
#gender
boxplot(affairs ~ gender, data = Affairs, main = "Number of Affairs by Gender", ylab = "Number of Affairs", xlab = "Gender")
wilcox.test(affairs ~ gender, data = Affairs)
```

We can determine that gender is not statistically significant, but with non-normal distribution these tests are not as reliable.

```{r}
# children
boxplot(affairs ~ children, data = Affairs, main = "Number of Affairs by Children", ylab = "Number of Affairs", xlab = "Children")
wilcox.test(affairs ~ children, data = Affairs)
```

With children, we can see it is statistically significant.

```{r}
library(GGally)
ggcorr(Affairs, label = T, method = c("all.obs","spearman"))
```

Most related vars to the target are: years married and rating of marital satisfaction

but are very weak

Check mean and variance of target

```{r}
mean<- mean(Affairs$affairs)
var <- var(Affairs$affairs)

var/mean
```

variance is 7 times bigger than the expected value of y. We can guess we have a problem of dispersion.

# split data

```{r}
training.samples <- Affairs$affairs %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- Affairs[training.samples, ]
test.data <- Affairs[-training.samples, ]
```

Simple Cross Validation 80 / 20

# Train and compare models

Let's train a linear and a poisson model to start with

```{r}
formula <- affairs ~ age + yearsmarried + religiousness + rating + rating*religiousness + children 
```

We are going to train a linear model to see how it compares to a Poisson regression.

```{r}
# Train a LM model
lm_model <- lm(formula, data = train.data)

summary(lm_model)

# Train a poisson model
poisson_model <- glm(formula, family = poisson, data = train.data)

poisson_model
```

From the linear model we see that rating and years married are the important variables.

This is the results of the poisson model. The residual deviance is lower than the null deviance, so it is better than a random model. However, it is much higher than the degrees of freedom, which means it does not fit the model.

Compare models using Achaiche information criteria (AIC)

```{r}
aic_results <- AIC(lm_model, poisson_model)

# Show results
list(AIC = aic_results)

```

Because our AIC is lower for the poisson model, we can see that the poisson is better than a linear model.

Lower AIC in Poisson -\> better model than linear for this data

Compare MAE for both models

```{r}
lm_preds <- predict(lm_model, type = "response", newdata = test.data)
poisson_preds <- predict(poisson_model, type = "response",  newdata = test.data)

lm_MAE <- mean(abs(test.data$affairs - lm_preds))
poisson_MAE <- mean(abs(test.data$affairs - poisson_preds))

lm_MAE
poisson_MAE
```

This is the residual of errors. Mean absolute error. Comparing them, neither are close to the correct answer of 1.5.

We can also tabulate predictions vs actual values like in a confusion matrix

```{r}
table(true = test.data$affairs, pred = round(poisson_preds))

```

Most people have zero numbers of affairs. But we are predicting a lot of values that should be zero. The poisson regression expects there to be more values than zeros, but in our data this is not the case.

```{r}
# Calculate Pearson residuals
fitted_values <- fitted(poisson_model)
pearson_residuals <- residuals(poisson_model, type = "pearson")

# Plot residuals vs fitted values
plot(fitted_values, pearson_residuals,
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```

We want to have a cloud of dots around the line, rather than patterns.

If residuals show a pattern, additional predictors or transformations may be needed.(Random scatter around zero suggests a well-fitting model)

One possible remedy is to incorporate interaction terms, capturing complex relationships between predictors.

This may improve model fit, but can lead to overfitting.

# Overdispersion -----

```{r}
# Extract residual deviance
residual_deviance <- poisson_model$deviance

# Extract residual degrees of freedom
residual_df <- poisson_model$df.residual

# Compute dispersion factor
dispersion_factor <- residual_deviance / residual_df

dispersion_factor
```

Quite high dispersion factor! The variance is very high in the model, and we have a problem with overdispersion. It should be lower than 1.5.

```{r}
# Overdispersion test
library(AER)
dispersiontest(poisson_model)
```

Ho: There is no overdispersion

Seems there is overdispersion (almost 0 p-value)

The p-value is very low, so we reject the null hypothesis.

# Quasipoisson and negative binomial alternatives ----

Quasi-poisson model

```{r}
# Train a poisson model
qpoisson_model <- glm(formula, family = quasipoisson, data = train.data)

qpoisson_model

# Compute predictions and MAE 
qpois_preds <-predict(qpoisson_model, type = "response", newdata = test.data)
qpois_MAE <- mean(abs(test.data$affairs - qpois_preds))

qpois_MAE

```

Same coefficients as Poisson, but corrected standard errors. AIC, BIC cannot be used with Quasi-Poisson regression.

Picking a new type of model. Quasi-poisson has a correction for the overdispersion.

The residual deviance has decreased a lot. The fit of the model is getting better, but the predictions will probably not get better because of the nature of the target variable.

Negative - binomial model

```{r}
# Negative binomial model 
library(MASS)
negbin_model<-glm.nb(formula, data = train.data)
negbin_model

# Compute predictions and MAE 
nb_preds <-predict(negbin_model, type = "response", newdata = test.data)
nb_MAE <- mean(abs(test.data$affairs - nb_preds))

nb_MAE
```

The residual deviance here is lowest, meaning the negative binomial is the best fit for this model.

# Zero-inflated models -----

ZI Poisson

```{r}
library(pscl)
zip <- zeroinfl(formula, data = train.data, dist="poisson")

zip_preds <- predict(zip, type = "response", newdata = test.data)
zip_mae <- mean(abs(test.data$affairs - zip_preds))

zip_mae
```

ZI Negative Binomial

```{r}
zi_nb <- zeroinfl(formula, data = train.data, dist="negbin")

zi_nb_preds<-predict(zi_nb, type = "response", newdata = test.data)
zinb_mae <- mean(abs(test.data$affairs - zi_nb_preds))

zinb_mae

```

# Compare models ----

In terms of MAE - we dont see much improvement in prediction accuracy

```{r}
list(
  LinearModel = list(MAE = lm_MAE),
  PoissonModel = list(MAE = poisson_MAE),
  QuasiPoisson = list(MAE = qpois_MAE),
  NegBin = list(MAE = nb_MAE),
  ZIP = list(MAE = zip_mae),
  ZI_NB = list(MAE =zinb_mae)
)
```

We can see there are no differences in the predictions. This is a very rare event that although the model is getting better, the predictions are staying the same.

Our target variable is what it is.

In terms of AIC - our models have been improving a lot!

```{r}
list(
  LinearModel = AIC(lm_model),
  PoissonModel = AIC(poisson_model),
 # QuasiPoisson = AIC(), AIC is not valid here
  NegBin = AIC(negbin_model),
  ZIP = AIC(zip),
  ZI_NB = AIC(zi_nb)
)
```

Better statistical fit (AIC) does not necessarily mean better prediction accuracy. Zero-inflated models describe the data-generating process better, especially the zeros, but they do not magically predict rare counts more accurately :(

Rare event prediction is inherently hard!!

# Using caret for cross validation and model comparison

Caret does not support zero inflated models, so we will go ahead with the rest.

```{r}
library(MASS)
library(caret)
library(e1071)


set.seed(123)
#CV setup: cv with 10 folds
train_control <- trainControl(method = "cv", number = 5)

#linear
lm_fit <- train(formula, data = train.data, 
                method = "lm", 
                trControl = train_control)

#poisson
poisson_fit <- train(formula, data = train.data, 
                     method = "glm", family = poisson(),
                     trControl = train_control)

#quasi poisson model (just change the family)
qpois_fit <- train(formula, data = train.data, 
               method = "glm", family=quasipoisson(),
               trControl = train_control)

#negative binomial 
neg_bin_fit <- train(formula, data = train.data, 
               method = "glm.nb",
               trControl = train_control)


```

Now we use resamples() to collect the results from our cv process into one object. It just collects results already computed inside each model and put it together.

```{r}
results <- resamples(list(linear = lm_fit, poisson = poisson_fit, qpois = qpois_fit, negbin = neg_bin_fit))
summary(results)

bwplot(results, metric = "MAE")

```

From cv training, it seems that negbin provides the best MAE but differences are very small!

Now we can compute our predictions and metrics on the *test set*, and select our final model.

# Regularization in GLMs

We've been talking about several different models for different types of problems.

Now that we know most models form the generalized linear models family, we can combine them with regularization methods to prevent overtiffing, handle multicollinearity, perform feature selection...

Let's see how to apply Lasso regularization to Poisson and Logistic Regression

# Remember Penalized Linear Model with caret

-   Caret supports a lot of different models from other packages (like glmnet) Because glmnet only supports: Gaussian / Binomial / Multinomial / Poisson distributions, other models cannot be included in this caret approach (like negbin)
-   We can customize the model fitting process by selecting:
    -   Cross validation method un *trainControl*
    -   Performance metrics (RMSE by default) in *metric*, we have to change them for classification
-   It includes an argument to preprocess data (scale)

EXAMPLE:

lambda \<- 10\^seq(-3, 2, length = 100)

lasso_lm \<- train( target \~., \# model formula data = train.data, method = "glmnet", \# training package trControl = trainControl("cv", number = 10), \# cv settings tuneGrid = expand.grid(alpha = 1, lambda = lambda), \# values for hyperparameters preProc = c("scale") \# standardization (key in regularization methods) )

# Penalized Logistic Regression ----

```{r}
data("PimaIndiansDiabetes2", package = "mlbench")

# remove rows with NA values
df <- na.omit(PimaIndiansDiabetes2)
```

Variables info:

-   pregnant: Number of times pregnant
-   glucose: Plasma glucose concentration (glucose tolerance test)
-   pressure: Diastolic blood pressure (mm Hg)
-   triceps: Triceps skin fold thickness (mm)
-   insulin: 2-Hour serum insulin (mu U/ml)
-   mass: Body mass index
-   pedigree: Diabetes pedigree function (metric of genetic predisposition)
-   age: Age (years)
-   diabetes: detected diabetes (yes or no) --\> target variable

```{r}
# vars description
str(PimaIndiansDiabetes2)

# summary
summary(PimaIndiansDiabetes2)
```

Lets expand our dataset by including all interactions and the squared terms of pregnant and age

```{r}
# model formula for expanded regressors
form <- diabetes ~ (pregnant + glucose  + pressure + triceps  + 
                   insulin  + mass     + pedigree + age)^2 +
                   I(pregnant^2) + I(age^2)
```

Because we are only testing the data we can do it in the formula, but if we are going to include it we would add it to the dataset.

## Data split

```{r}
training.samples <- df$diabetes %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- df[training.samples, ]
test.data <- df[-training.samples, ]
```

## Train Lasso, Rigde and standard logistic regression with *caret*

Lambda possible values

```{r}
lambda <- 10^seq(-3, 1, length = 100)
```

Setting for cross validation, adjusted for classification

```{r}
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

### Standard Logistic

```{r}
logistic <- train(
 form, 
  data = train.data, 
  method = "glm", family = "binomial",
  trControl = ctrl,
  metric = "ROC", # we want to use the ROC function to select the best model
  preProc = c("scale")
 ) 

logistic
```

ROC, Sens, Spec are the metrics we get in the typical regression.

Why do we create a baseline model if we are going to do regularization? It is good to see how the regularization will change the model.

### Lasso

```{r}
logistic_lasso <- train(
 form, 
  data = train.data, 
  method = "glmnet", family = "binomial",
  trControl = ctrl,
  tuneGrid = expand.grid(alpha = 1, lambda = lambda),
  metric = "ROC",
  preProc = c("scale") # preprocessing step to scale numeric variables
 ) 

logistic_lasso
```

We get the logistical lasso estimation. This is a summary of the things that are tried during the cross-validation. It is changing the lambda value to try various combinations.

```{r}
# Get the best model's coefficients
lasso_best_lambda <- logistic_lasso$bestTune$lambda
lasso_coefficients <- coef(logistic_lasso$finalModel, s = lasso_best_lambda)

lasso_coefficients
```

This gives us the best combination, which was 0.004. When we use this lambda, we get a distinct set of coefficients. Lasso removed a lot of variables.

Note: Remember that regularized coefficients are best interpreted in terms of prediction, and variable selection — not direct impact on the target on its original scale

If we want to *interpret*, we'd better compute the standard logistic model with selected variables. Here we would interpret coefficients normally in a logistic regression

### Ridge

```{r}
logistic_ridge <- train(
 form, 
  data = train.data, 
  method = "glmnet", family = "binomial",
  trControl = ctrl,
  tuneGrid = expand.grid(alpha = 0, lambda = lambda),
  metric = "ROC",
  preProc = c("scale")
 ) 

logistic_ridge

logistic_ridge$bestTune$lambda
```

This ROC is a little higher than the ones we got before. The best lambda is 0.02.

# Create predictions and evaluate models on test data

Predicted probabilities

```{r}
logit_prob <- predict(logistic, test.data, type= "prob")[, "pos"]
logit_lasso_prob <- predict(logistic_lasso, test.data, type= "prob")[, "pos"]
logit_ridge_prob <- predict(logistic_ridge, test.data, type= "prob")[, "pos"]
```

AUC

```{r}
library(pROC)

auc_logit <- roc(test.data$diabetes, logit_prob,  levels = c("neg","pos"))$auc
auc_lasso <- roc(test.data$diabetes, logit_lasso_prob,  levels = c("neg","pos"))$auc
auc_ridge <- roc(test.data$diabetes, logit_ridge_prob,  levels = c("neg","pos"))$auc

```

Comparing the three models.

```{r}
results <- data.frame(
  Model   = c("Logistic", "Ridge", "LASSO"),
  AUC      = c(auc_logit, auc_ridge, auc_lasso))

results

```

In this case, Ridge achieves the best predictive performance on the test set, probably because predictors are correlated and most of them contain useful information for the target

Ridge ROC and AUC plot

```{r}
plot.roc(test.data$diabetes, logit_ridge_prob, 
         col="darkblue", # curve color
         print.auc = TRUE,  # add auc value
         auc.polygon=TRUE, # add auc shape
         max.auc.polygon=TRUE,
         auc.polygon.col="lightblue", # auc color
         print.thres="best") # best thresholds
```

# Penalized Poisson -----

```{r}
data(Affairs, package = 'AER')
```

Infidelity data, known as Fair's Affairs. Cross-section data from a survey conducted by Psychology Today in 1969.

-   affairs: is the number of extramarital affairs in the past year and is our response variable
-   gender
-   age
-   yearsmarried
-   children (yes or no)
-   religiousness (1 - anti, 5 - very)
-   education
-   rating self rating of the marriage 1 (very unhappy) to 5 (very happy)

```{r}
summary(Affairs)
```

# EDA

```{r}
library(DataExplorer)

plot_intro(Affairs)
```

Check distributions

```{r}
#histogram
hist(Affairs$affairs)

#mean and variance 
mean(Affairs$affairs)
var(Affairs$affairs)

#density plot 
success <- min(Affairs$affairs):max(Affairs$affairs)

plot(density(Affairs$affairs),ylim=c(0,0.5), xlim = c(0,15))+
lines(success,dpois(success, lambda=mean(Affairs$affairs)),col="green")
```

```{r}
library(GGally)
ggcorr(Affairs, label = T, method = c("all.obs","spearman"))
```

## Feature engineering

Fix some variables and expand the dataset

```{r}
Affairs <- Affairs %>% mutate(gender = ifelse(gender == "male", 1,0),
                   children = ifelse(children == "yes",1,0))

Affairs$occupation <- as.factor(Affairs$occupation )

Affairs <- Affairs %>% mutate(rat_rel = rating*religiousness,
                              rating2 = rating^2,
                              yearsmarried2 = yearsmarried^2, 
                              age2 = age^2)



```

# Split data

```{r}
set.seed(123)
training.samples <- Affairs$affairs %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- Affairs[training.samples, ]
test.data <- Affairs[-training.samples, ]

```

# Modeling

CV settings

```{r}
ctrl <- trainControl(
  method = "cv",
  number = 5
)
```

Model formula

```{r}
form <- affairs ~ .
```

# Let's perform regulaization in Poisson model

### Standard poisson (not regularized) as a baseline

```{r}
set.seed(123)

poisson_fit <- train(
  form,
  data = train.data,
  method = "glm", family = "poisson",
  trControl = ctrl,
  preProcess = c("scale")
)

poisson_fit

```

### Lasso poisson

```{r}
set.seed(123)
lasso_fit <- train(
  form,
  data = train.data,
  method = "glmnet",
  family = poisson,
  trControl = ctrl,
  preProcess = c("scale"),
  tuneGrid = expand.grid(
    alpha = 1, # LASSO
    lambda = 10^seq(-3, 2, length = 100)
  )
)

lasso_fit
```

```{r}
coef(lasso_fit$finalModel, s = lasso_fit$bestTune$lambda)

```

Lasso removed all the variables from the dataset.

# Ridge

```{r}
set.seed(123)
ridge_fit <- train(
  form,
  data = train.data,
  method = "glmnet",
  family = poisson,
  trControl = ctrl,
  preProcess = c("scale"),
  tuneGrid = expand.grid(
    alpha = 0,# Ridge
    lambda = 10^seq(-3, 2, length = 50)
  )
)

ridge_fit
plot(ridge_fit)

```

Lets see regularization applied in each case!

```{r}
lasso_fit$bestTune
ridge_fit$bestTune

```

The best lambda in both cases is lambda = 100

# Evaluate the models on test data

```{r}
#| eval: false
pred_lasso <- predict(lasso_fit, newdata = test.data)
pred_ridge <- predict(ridge_fit, newdata = test.data)
pred_poisson <- predict(poisson_fit, newdata = test.data)

data.frame(
  Model = c("Poisson", "LASSO", "Ridge"),
  MAE   = c(
    MAE(pred_lm, test.data$affairs),
    MAE(pred_lasso, test.data$affairs),
    MAE(pred_ridge, test.data$affairs)
  ),
  RMSE  = c(
    RMSE(pred_lm, test.data$affairs),
    RMSE(pred_lasso, test.data$affairs),
    RMSE(pred_ridge, test.data$affairs)
  )
)

```

Regularization protects against overfitting only when overfitting is a "danger". In this case, even with the engineered features, predictors barely explain the count variability (most observations are just 0).

If the underlying data pattern is weak (as we saw in our previous example with this data), regularization often does not improve or even hurts predictive performance. LASSO and Ridge, by shrinking coefficients, are actually throwing away small but real effects — because they can’t distinguish them from noise in such a weak-signal setting.
