---
title: "Survey Methodology"
format: html
editor: visual
---

# Packages

```{r}
library(tidyverse) # for tidyverse
library(here) # for file paths
library(survey) # for survey analysis
library(srvyr) # for tidy survey analysis
library(sampling)
```

# Tidy Survey Analysis in R using the srvyr Package

## Goals

Grading component next week. He will show us code and we will do exercises that we have to hand in at the end of class.

At the end of this workshop series, we will be able to:

-   specify a survey design in R to create a survey object

-   calculate point estimates and their standard errors with survey data

    -   means, quantiles, ratios

    -   Proportions, totals, and counts

-   Perform t-tests

-   Fit regression models

## Resources

[Survey Sampling Package](https://cran.r-project.org/web/packages/sampling/index.html)

Stephanie A. Zimmer, Rebecca J. Powell, and Isabella C. Velásquez. 2024. Exploring Complex Survey DataAnalysis Using R. CRC Press

## Sampling Review

-   **Simple random sampling:** every unit has the same chance of being selected

    -   Without replacement: units can only be selected once

    -   With replacement: units can be selected more than once

-   **Systematic sampling:** sample of n individuals from an ordered list and sampling individuals at an interval with a random starting point

-   **Probability proportional to size:** probability of selection is proportional to "size"

-   **Stratified sampling:** divide population into mutually exclusive subgroups (strata). Randomly sample within each stratum

-   **Clustered sampling:** divide population into mutually exclusive subgroups (clusters). Randomly sample clusters and then individuals within clusters

    -   implemented because of practical concerns

[**Important Note:**]{.underline}

Simple Random Sampling is not really achievable, but it is what our statistical metrics are based on.

## Package

The sampling package provides a set of useful functions for...

-   [Sampling]{.underline}: Stratification, two-stage, unequal probabilities, balanced sampling

-   [Estimation]{.underline}: calibration and regression estimator

-   [Tools]{.underline}: computation of inclusion probabilities, crossing strata

-   [Data bases]{.underline}: Swiss municipalities, Belgian municipalities.

### Example Use Case: select a sample size of 10 out of a population of 50

```{r}
s <- srswor(n = 10,## Sample size
            N = 50 ## Population size
)
matrix(s, nrow = 5, byrow = TRUE)

# selected example ids
which(s == 1)
```

### Example Use Case: Stratified Sampling

```{r}
# variable of stratification in this case 3
group <- rep(1:3, each = 5)

# matrix of balancing variables
X <- cbind(1:15)

# vector of inclusion probabilities (sample of 9)
pik <- rep(3/5, times = 15)

# selection of a stratified sample
s <- balancedstratification(X = X, strata = group, pik = pik, comment = FALSE)

# the sample is
matrix(s, nrow = 3, byrow = TRUE)
which(s==1)
```

### Example Use Case: Clustered Sampling

```{r}
# selection of 2 clusters
s <- balancedcluster(
   X = X, m = 2, cluster = group,
   selection = 2, # cluster selection equal or by size
   comment = FALSE
)

# sample of clusters with the inclusion probabilities of the clusters
head(s)

# selected clusters
unique(group[s[,1]==1])

# selected units
which(s[,1]==1)
```

# Overview of Survey Analysis using srvyr Package

1.  Create a `tbl_svy` object using: `as_survey_design` or `as_survey_rep`
2.  Subset data (if needed) using `filter` (sub-populations)
3.  Specify domains of analysis using `group_by`
4.  Within `summarize`, specify variables to calculate including means, totals, proportions, quantiles, and more

If we do clustered random sampling, we have to correct the point estimates. Let's say we know education has a bell-shaped distribution. Some clusters are rich and some are poor, so the distribution might be skewed and not match the population distribution.

When we compute standard errors, we are assuming simple random sampling. A simple random sample would be more precise. There is less variation you are covering in your clusters than you have in the overall population.

If we don't take into the account the sampling strategy, we will underestimate the uncertainty. Our confidence intervals are tighter than they should be.

## How to Determine the Design

1.  Look at the documentation associated with the analysis file
2.  keywords to look for: methodology, design, analysis guide, technical documentation
3.  documentation will indicate the variables needed to specify the design. Look for:
    -   weight (almost always)
    -   strata, clusters, PSUs; OR pseudo-strata, pseudo-cluster (a method to anonymitize the data); OR replicate weights (a method to anonymitize the data)
    -   finite population correction or population sizes
4.  Documentation may include syntax for SAS, SUDAAN, Stata, and/or R

Weight is the inverse of the inclusion probability.

### Surveys

-   [American National Election Study (ANES) 2020](https://electionstudies.org/data-center/2020-time-series-study/)

    -   Example of weighting

-   [Residential Energy Consumption Survey (RECS) 2015](https://www.eia.gov/consumption/residential/index.php)

    -   Example of replicate weights

```{r}
library(survey)
RECS15 <- read.csv(
  file = './Data/recs.rds',
  header=TRUE,
  sep= ","
)
sampweights <- RECS15$NWEIGHT
brrwts <- RECS15[, grepl(“^BRRWT”, names(RECS15))]
des <- svrepdesign(
  weights = sampweights,
  repweights = brrwts,
  type = "Fay",
  rho = 0.5,
  mse = TRUE,
  data = RECS15)
```

## Weighted Analysis for Continuous Variables: Domain Analysis

If we want to get estmates by another variable, we need to add a `group_by` statement before doing the analysis.

Example: What is the average amount of dollar spent on electricity for households that use AC and those that do not use AC?

```{r}
recs_des |> 
  group_by(ACUsed) |> 
  summarize( ElBill=survey_mean(DOLLAREL, na.rm=TRUE) )
```
