---
title: "Topic 3.3: Tree-Based Models"
subtitle: "Decision Trees, Bagging, and Random Forests in R"
author: "Advanced Modelling, Master in Computational Social Science"
date: 'UC3M, 2026'
output:
  html_document:
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(3301)
```

# Introduction

In this notebook we will practice with the tree-based methods seen in topic 3.3.

-   **Decision Trees**: simple, interpretable binary decision rules
-   **Pruning**: controlling overfitting by simplifying the tree
-   **Bagging**: averaging many trees grown on bootstrap resamples
-   **Random Forests**: bagging plus random feature subsampling at each split

We will start with some small toy examples (`iris` and `mtcars`) to see what are the main R functions and their arguments, and then apply the full workflow to our real case study of student failure in mathematics as we did in topic 3.1.

```{r libraries}
library(tidyverse)
library(tree)          # tree() and cv.tree() for decision trees
library(randomForest)  # randomForest() for bagging and random forests
library(caret)         # createDataPartition(), confusionMatrix()
```

------------------------------------------------------------------------

# Classification Tree (`iris`)

We use the classic Iris dataset (3 flower species, 4 numeric predictors) to learn the basic tree workflow. We start by creating a stratified train/test split with `caret`, and in this first example we focus on `Petal.Length` and `Petal.Width` so we can visualise the tree partition directly in 2D.

```{r iris-split}
data(iris)

set.seed(123)
iris_idx <- createDataPartition(iris$Species, p = 0.7, list = FALSE)
iris_train <- iris[iris_idx, ]
iris_test  <- iris[-iris_idx, ]

cat("Train:", nrow(iris_train), "| Test:", nrow(iris_test), "\n")
```

The main function is `tree()` from the `tree` package. Some of its arguments are:

-   `formula`
-   `data`
-   `control = tree.control(...)` which governs how large the tree grows with the following parameters:
    -   `nobs` (number of observations, required)
    -   `mindev` (minimum deviance improvement required to attempt a split; smaller values grow bigger trees, default `0.01`)
    -   `minsize` (minimum observations per terminal node, default `10`).

```{r iris-tree-fit}
iris_tree <- tree(
  Species ~ Petal.Length + Petal.Width,
  data    = iris_train,
  control = tree.control(
    nobs    = nrow(iris_train),
    mindev  = 0.01,   # lower = bigger tree
    minsize = 5       # min obs per leaf
  )
)

summary(iris_tree)
```

The summary shows the number of terminal nodes, the variables actually used in splits, and the **training** misclassification rate. We can visualise the full tree with `plot()` and `text()`. `pretty = 0` keeps the original factor labels rather than abbreviating them.

```{r iris-tree-plot, fig.height=6}
plot(iris_tree)
text(iris_tree, pretty = 0, cex = 0.8)
title("Iris classification tree")
```

Each internal node shows the splitting variable and threshold; each leaf shows the predicted class.

Because we used two predictors, we can plot the feature-space partition induced by the tree.

```{r iris-tree-partition, fig.height=5}
iris_grid <- expand.grid(
  Petal.Length = seq(min(iris_train$Petal.Length) - 0.3,
                     max(iris_train$Petal.Length) + 0.3,
                     length.out = 200),
  Petal.Width  = seq(min(iris_train$Petal.Width) - 0.3,
                     max(iris_train$Petal.Width) + 0.3,
                     length.out = 200)
)

iris_grid$Pred <- predict(iris_tree, newdata = iris_grid, type = "class")

ggplot() +
  geom_tile(
    data = iris_grid,
    aes(x = Petal.Length, y = Petal.Width, fill = Pred),
    alpha = 0.28
  ) +
  geom_point(
    data = iris_train,
    aes(x = Petal.Length, y = Petal.Width, color = Species),
    size = 1.6,
    alpha = 0.85
  ) +
  labs(
    title = "Iris tree partition in petal feature space",
    subtitle = "Background: class predicted by the tree",
    x = "Petal.Length",
    y = "Petal.Width"
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal()
```

To evaluate on the test set we use `predict()` with `type = "class"`, which returns the predicted class directly instead of the default probability matrix.

```{r iris-tree-eval}
iris_tree_pred <- predict(iris_tree, newdata = iris_test, type = "class")
confusionMatrix(iris_tree_pred, iris_test$Species)
```

A large tree may overfit. Pruning addresses this by first growing a full tree and then using cross-validation to select the number of leaves that minimises CV error. The workflow is:

1.  Grow a large tree with `mindev = 0`
2.  Run `cv.tree()` with `FUN = prune.misclass` to use misclassification error instead of deviance
3.  Pick the size with the lowest CV error
4.  Call `prune.misclass(big_tree, best = chosen_size)`.

The `cv.tree` object contains `$size` (number of terminal nodes), `$dev` (CV error), and `$k` (the regularisation penalty alpha).

```{r iris-big}
iris_big <- tree(
  Species ~ .,
  data    = iris_train,
  control = tree.control(nobs = nrow(iris_train), mindev = 0, minsize = 2)
)

plot(iris_big)
text(iris_big, pretty = 0, cex = 0.85)
title(paste("Original Big Tree"))
```

```{r iris-pruning}

set.seed(3302)
iris_cv <- cv.tree(iris_big, FUN = prune.misclass)

# Show the results
iris_cv

# Create a plot
tibble(size = iris_cv$size, cv_error = iris_cv$dev) %>%
  ggplot(aes(x = size, y = cv_error)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Pruning curve: CV misclassification vs tree size",
    x = "Number of terminal nodes",
    y = "CV misclassification count"
  ) +
  theme_minimal()
```

```{r iris-choosing-min}
# Selecting the tree size with minimum CV error
best_size <- iris_cv$size[which.min(iris_cv$dev)]
cat("Best tree size (terminal nodes):", best_size, "\n")

# But this is the same size as the big one!
# In this case there are several values with minimum CV error
# We should then choose the smallest size one given that they have the same CV error
# The following code does precisely that
min_dev <- min(iris_cv$dev, na.rm = TRUE)
best_size <- min(iris_cv$size[iris_cv$dev == min_dev], na.rm = TRUE)
cat("Best tree size (terminal nodes):", best_size, "\n")
```

```{r iris-pruned-eval}
iris_pruned <- prune.misclass(iris_big, best = best_size)

plot(iris_pruned)
text(iris_pruned, pretty = 0, cex = 0.85)
title(paste("Pruned tree ", best_size, "leaves"))

iris_pruned_pred <- predict(iris_pruned, newdata = iris_test, type = "class")
confusionMatrix(iris_pruned_pred, iris_test$Species)
```

You can see how the accuracy is the same as the unprunned tree, but its smaller in size and therefore more interpretable.

------------------------------------------------------------------------

# Bagging and Random Forest (`iris`)

To use bagging, we saw that its indeed an special case of a random forest where `m = p`. Therefore both methods come from the `randomForest` package and share the same interface, where the key argument is `mtry`, the number of candidate predictors randomly sampled at each split. Setting `mtry = p` (all predictors) gives **bagging**, while `mtry = floor(sqrt(p))` (the default for classification) gives a **random forest**. The argument `ntree` controls the total number of trees (default 500; more trees are rarely harmful as it will not overfit due to the averaging, just slower), and `importance = TRUE` enables variable importance computation.

Starting with bagging we will fit 500 trees, each seeing all 4 predictors at every split (`m=p`).

```{r iris-bagging}
p <- ncol(iris_train) - 1   # number of predictors

set.seed(3303)
iris_bag <- randomForest(
  Species   ~ .,
  data       = iris_train,
  mtry       = p,           # bagging: use ALL predictors at each split
  ntree      = 500,
  importance = TRUE
)

iris_bag   # prints OOB confusion matrix and error
```

Remember that the **OOB (out-of-bag) error** printed by the model is a built-in estimate of test error at no extra cost: each tree only uses \~63 % of training cases, so the remaining \~37 % act as a held-out set for that tree. Therefore we dont have the need to explicitly use cross-validation

```{r iris-bagging-eval}
iris_bag_pred <- predict(iris_bag, newdata = iris_test, type = "class")
confusionMatrix(iris_bag_pred, iris_test$Species)
```

Now we fit the random forest, reducing `mtry` to `floor(sqrt(p))` so that only a random subset of features is considered at each split. This decorrelates the trees and typically improves the ensemble.

```{r iris-rf}
set.seed(3304)
iris_rf <- randomForest(
  Species   ~ .,
  data      = iris_train,
  mtry      = floor(sqrt(p)),  # default: sqrt(p) for classification
  ntree     = 500,
  importance = TRUE
)

iris_rf
```

```{r iris-rf-eval}
iris_rf_pred <- predict(iris_rf, newdata = iris_test, type = "class")
confusionMatrix(iris_rf_pred, iris_test$Species)
```

In this case, as `iris` is a very simple dataset we get the same accuracy with all the models, but we will later see how this changes in more complex datasets.

With `importance = TRUE` we can inspect which variables mattered most. `importance()` returns two types of scores: **MeanDecreaseAccuracy** (`type = 1`) measures how much accuracy drops when each variable is permuted at random (higher = more important), and **MeanDecreaseGini** (`type = 2`) sums the total node impurity decrease from splits on that variable across all trees. `varImpPlot()` visualises both at once.

```{r iris-importance}
varImpPlot(iris_rf, main = "Iris Random Forest — Variable Importance", cex = 0.9)
```

------------------------------------------------------------------------

# Regression Tree (mtcars)

We have been using tree-based models with classification examples but trees also work for **continuous targets**. The syntax is identical: `predict()` then returns a numeric vector and the tree minimises residual sum of squares (RSS) at each split instead of impurity. We use `mtcars` to predict fuel efficiency (`mpg`) from the other car features.

```{r mtcars-split}
data(mtcars)
set.seed(456)
mt_idx   <- createDataPartition(mtcars$mpg, p = 0.75, list = FALSE)
mt_train <- mtcars[mt_idx, ]
mt_test  <- mtcars[-mt_idx, ]
```

```{r mtcars-tree-fit}
mt_tree <- tree(mpg ~ ., data = mt_train)

plot(mt_tree)
text(mt_tree, cex = 0.75)
title("mtcars regression tree")
```

To see clearly how regression trees create piecewise-constant predictions, we fit a one-predictor tree (`mpg ~ wt`) and draw its fitted values as a step function.

```{r mtcars-tree-step, fig.height=5}
mt_tree_wt <- tree(mpg ~ wt, data = mt_train)

wt_grid <- tibble(
  wt = seq(min(mt_train$wt), max(mt_train$wt), length.out = 300)
) %>%
  mutate(mpg_hat = predict(mt_tree_wt, newdata = .))

ggplot(mt_train, aes(x = wt, y = mpg)) +
  geom_point(color = "grey35", alpha = 0.75) +
  geom_step(
    data = wt_grid,
    aes(x = wt, y = mpg_hat),
    color = "firebrick",
    linewidth = 1
  ) +
  labs(
    title = "Regression tree as a step function (mpg ~ wt)",
    subtitle = "Each flat segment is one terminal-node mean",
    x = "wt (1000 lbs)",
    y = "mpg"
  ) +
  theme_minimal()
```

```{r mtcars-tree-eval}
mt_pred <- predict(mt_tree, newdata = mt_test)

rmse <- sqrt(mean((mt_pred - mt_test$mpg)^2))
cat("Test RMSE:", round(rmse, 3), "\n")

tibble(actual = mt_test$mpg, predicted = mt_pred) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(size = 2, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Regression tree: predicted vs actual mpg") +
  theme_minimal()
```

------------------------------------------------------------------------

# Real Case Study: Student Failure

Now we apply the full workflow to the student failure dataset introduced in topic 3.1. The goal is to predict whether a student **fails mathematics** (final grade `G3 < 5`). We work with the same compact feature set used before: `failures`, `absences`, `studytime`, `Medu`, `Fedu`, `Walc`, `goout`, and `age`. Please refer to the topic 3.1 notebook for the full variable descriptions and background.

```{r student-data}
math_raw <- read.csv2("student-mat.csv",
  stringsAsFactors = TRUE
)

dim(math_raw)
summary(math_raw$G3)
```

```{r student-preprocess}
math <- math_raw %>%
  mutate(Fail = factor(G3 < 5, levels = c(FALSE, TRUE), labels = c("No", "Yes"))) %>%
  dplyr::select(Fail, failures, absences, studytime, Medu, Fedu, Walc, goout, age)

summary(math)
prop.table(table(math$Fail))
```

Remember that the dataset is imbalanced: failures are the minority class (\~17–20 %). We should keep this in mind when reading the metrics.

We do a stratified 80/20 split to preserve class proportions in both sets.

```{r student-split}
set.seed(2026)
in_train <- createDataPartition(math$Fail, p = 0.8, list = FALSE)
train_df <- math[in_train, ]
test_df  <- math[-in_train, ]

cat("Train:", nrow(train_df), "| Test:", nrow(test_df), "\n")
cat("Train failure rate:", round(mean(train_df$Fail == "Yes"), 3), "\n")
cat("Test  failure rate:", round(mean(test_df$Fail  == "Yes"), 3), "\n")
```

We fit a decision tree first. The tree will automatically select which variables to split on and in what order.

```{r student-tree}
student_tree <- tree(
  Fail ~ .,
  data    = train_df,
  control = tree.control(nobs = nrow(train_df), mindev = 0.01, minsize = 10)
)

summary(student_tree)
```

```{r student-tree-plot, fig.height=7}
plot(student_tree)
text(student_tree, pretty = 0, cex = 0.75)
title("Student failure — decision tree")
```

```{r student-tree-eval}
student_tree_pred <- predict(student_tree, newdata = test_df, type = "class")
cm_tree <- confusionMatrix(student_tree_pred, test_df$Fail, positive = "Yes")
cm_tree
```

To guard against overfitting we run the same pruning workflow: grow a maximal tree, cross-validate to find the optimal size, then prune back.

```{r student-pruning}
student_big <- tree(
  Fail ~ .,
  data    = train_df,
  control = tree.control(nobs = nrow(train_df), mindev = 0, minsize = 2)
)

set.seed(3305)
student_cv <- cv.tree(student_big, FUN = prune.misclass)

tibble(size = student_cv$size, cv_error = student_cv$dev) %>%
  ggplot(aes(x = size, y = cv_error)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_vline(
    xintercept = student_cv$size[which.min(student_cv$dev)],
    linetype = "dashed", color = "firebrick"
  ) +
  labs(
    title = "Pruning curve: CV misclassification vs tree size",
    x = "Terminal nodes", y = "CV misclassification count"
  ) +
  theme_minimal()
```

```{r student-pruned}
best_size <- student_cv$size[which.min(student_cv$dev)]
cat("Best size:", best_size, "\n")

student_pruned      <- prune.misclass(student_big, best = best_size)
student_pruned_pred <- predict(student_pruned, newdata = test_df, type = "class")
cm_pruned <- confusionMatrix(student_pruned_pred, test_df$Fail, positive = "Yes")
cm_pruned
```

We now move to ensemble methods. For bagging we set `mtry = p` so all predictors are candidates at every split; for the random forest we use `mtry = floor(sqrt(p))` to decorrelate trees. Both models print the OOB error matrix directly.

```{r student-bagging}
p_student <- ncol(train_df) - 1

set.seed(3306)
student_bag <- randomForest(
  Fail       ~ .,
  data       = train_df,
  mtry       = p_student,   # bagging: all predictors
  ntree      = 500,
  importance = TRUE
)

student_bag
```

```{r student-bagging-eval}
student_bag_pred <- predict(student_bag, newdata = test_df, type = "class")
cm_bag <- confusionMatrix(student_bag_pred, test_df$Fail, positive = "Yes")
cm_bag
```

```{r student-rf}
set.seed(3307)
student_rf <- randomForest(
  Fail       ~ .,
  data       = train_df,
  mtry       = floor(sqrt(p_student)),
  ntree      = 500,
  importance = TRUE
)

student_rf
```

```{r student-rf-eval}
student_rf_pred <- predict(student_rf, newdata = test_df, type = "class")
cm_rf <- confusionMatrix(student_rf_pred, test_df$Fail, positive = "Yes")
cm_rf
```

A useful diagnostic for the random forest is to plot the OOB error as a function of the number of trees, to check whether 500 trees was sufficient for the estimate to stabilise.

```{r student-oob-curve}
tibble(
  trees     = seq_len(nrow(student_rf$err.rate)),
  oob_error = student_rf$err.rate[, "OOB"]
) %>%
  ggplot(aes(x = trees, y = oob_error)) +
  geom_line(color = "darkorange", linewidth = 1) +
  labs(
    title    = "Random Forest OOB error curve",
    subtitle = "Error stabilises after a few hundred trees",
    x = "Number of trees", y = "OOB error rate"
  ) +
  theme_minimal()
```

With all four models fitted we can now compare their test-set performance in one table. Because the classes are imbalanced, we report **Balanced Accuracy** alongside raw accuracy — a model that always predicts "No failure" would score high accuracy but zero sensitivity.

```{r student-comparison}
extract_metrics <- function(cm, model_name) {
  tibble(
    Model               = model_name,
    Accuracy            = as.numeric(cm$overall["Accuracy"]),
    Sensitivity         = as.numeric(cm$byClass["Sensitivity"]),
    Specificity         = as.numeric(cm$byClass["Specificity"]),
    `Balanced Accuracy` = as.numeric(cm$byClass["Balanced Accuracy"])
  )
}

results <- bind_rows(
  extract_metrics(cm_tree,   "Decision tree"),
  extract_metrics(cm_pruned, "Pruned tree"),
  extract_metrics(cm_bag,    "Bagging"),
  extract_metrics(cm_rf,     "Random Forest")
)

knitr::kable(results, digits = 3)
```

```{r student-comparison-plot}
results %>%
  dplyr::select(Model, Accuracy, `Balanced Accuracy`, Sensitivity) %>%
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = reorder(Model, Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  ylim(0, 1) +
  labs(title = "Student failure: model comparison", x = "", y = "Score") +
  theme_minimal()
```

Finally, we inspect variable importance from the random forest. `varImpPlot()` shows both the permutation-based score (MeanDecreaseAccuracy, `type = 1`) and the Gini-based score (MeanDecreaseGini, `type = 2`). Note that importance ranks variables by predictive contribution, not causal effect — correlated variables can share or swap importance values.

```{r student-importance}
varImpPlot(
  student_rf,
  main = "Student failure RF — Variable Importance"
)
```

```{r student-importance-table}
importance(student_rf, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("Variable") %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  knitr::kable(digits = 3)
```

------------------------------------------------------------------------

# What about predicting probabilities?

What about setting a different threshold? Remember what we saw about imbalanced data in classification, where the problem can be partly fixed by choosing a threshold different from 0.5.

For the `tree` package, class probabilities are obtained with `predict(..., type = "vector")` (while `type = "class"` returns only the hard class). Below we reuse the first Iris tree example and compare both outputs.

```{r iris-probabilities}
# Hard class predictions from the first iris tree section
iris_tree_class <- predict(iris_tree, newdata = iris_test, type = "class")

# Probability matrix (one column per species)
iris_tree_prob <- predict(iris_tree, newdata = iris_test, type = "vector")

# Recover class labels from probabilities via argmax
iris_tree_from_prob <- colnames(iris_tree_prob)[max.col(iris_tree_prob, ties.method = "first")]
iris_tree_from_prob <- factor(iris_tree_from_prob, levels = levels(iris_test$Species))

cat("Hard class == argmax(probabilities):", all(iris_tree_class == iris_tree_from_prob), "\n")

bind_cols(
  tibble(true_species = iris_test$Species, pred_class = iris_tree_class),
  as_tibble(iris_tree_prob) %>% rename_with(~ paste0("p_", .x))
) %>%
  # slice_head(n = 10) %>%
  knitr::kable(digits = 3)
```

------------------------------------------------------------------------

# Some R packages for tree-based models

-   **`tree`**: lightweight package for classification and regression trees with cross-validation pruning.
-   **`rpart`** + **`rpart.plot`**: more feature-rich alternative for decision trees with nicer visualisation.
-   **`randomForest`**: the classic implementation of bagging and random forests.
-   **`ranger`**: faster, parallelised random forest with the same interface as `randomForest`.
-   **`xgboost`** / **`lightgbm`**: gradient boosting: a more advanced ensemble strategy.
-   **`caret`** / **`tidymodels`**: unified interfaces that wrap all the above for cross-validation, hyperparameter tuning, and pipelines.

# Practice suggestion

Try to replicate the study with the Portugese subject dataset:

```{r practice-setup, eval = FALSE}
por_raw <- read.csv2("student-por.csv",
  stringsAsFactors = TRUE
)

por <- por_raw %>%
  mutate(Fail = factor(G3 < 5, levels = c(FALSE, TRUE), labels = c("No", "Yes"))) %>%
  dplyr::select(Fail, failures, absences, studytime, Medu, Fedu, Walc, goout, age)

summary(por)
prop.table(table(por$Fail))
```
