---
title: "Topic 3.5: Neural Networks and MLPs"
subtitle: "Binary classification, multiclass classification, and regression in R"
author: "Advanced Modelling, Master in Computational Social Science"
date: "UC3M, 2025-2026"
output:
  html_document:
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In this notebook we practice the basic workflow for fitting a simpel MLPs network in R, using `caret` and `nnet` packages.

> Have a look at the `nnet` function documentation: [Link](https://www.rdocumentation.org/packages/nnet/versions/7.3-20/topics/nnet)

There are many other options to train NNs in R, such as the `tensorflow` package (which requires installing Python in the background) or the `torch` package, but these are out of the scope of this course.

We will solve 3 example toy problems:

1. binary classification on `iris` modified to be 2 classes
2. multiclass classification on full `iris`
3. regression on Sacramento house prices



```{r libraries}
library(tidyverse)
library(caret)
library(nnet)

theme_set(theme_minimal())
```

# Binary Classification on `iris`

We start with the lightest possible example: turn `iris` into a binary problem by asking whether a flower is `setosa` or not.

To keep the geometry intuitive, we only use `Petal.Length` and `Petal.Width` in this first example.

```{r iris-binary-data}
data(iris)

iris_binary <- iris %>%
  transmute(
    Petal.Length,
    Petal.Width,
    Class = factor(
      if_else(Species == "setosa", "setosa", "other"),
      levels = c("setosa", "other")
    )
  )

set.seed(3501)
iris_binary_idx <- createDataPartition(iris_binary$Class, p = 0.7, list = FALSE)
iris_binary_train <- iris_binary[iris_binary_idx, ]
iris_binary_test  <- iris_binary[-iris_binary_idx, ]

cat("Train:", nrow(iris_binary_train), "| Test:", nrow(iris_binary_test), "\n")
```

Before fitting the model, it helps to visualize the two classes:

```{r iris-binary-plot}
ggplot(iris_binary_train, aes(x = Petal.Length, y = Petal.Width, color = Class)) +
  geom_point(size = 2, alpha = 0.85) +
  labs(
    title = "Binary iris task: setosa vs other",
    subtitle = "We only use two petal features in this first example"
  ) +
  scale_color_brewer(palette = "Set1")
```


For this first model we use a small tuning grid:

- `size`: number of hidden units,
- `decay`: weight decay, a simple regularization parameter.

`caret` handles the tuning and control

```{r iris-binary-fit}
ctrl_class <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 1,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(3502)
iris_binary_fit <- train(
  Class ~ .,
  data = iris_binary_train,
  method = "nnet",
  metric = "Accuracy",
  preProcess = c("center", "scale"),
  trControl = ctrl_class,
  tuneGrid = expand.grid(
    size = c(1, 3, 5),
    decay = c(0, 0.01, 0.1)
  ),
  trace = TRUE,
  maxit = 300
)

iris_binary_fit
```


For binary classification, one very useful output is the predicted probability of the positive class.

```{r iris-binary-probabilities}
iris_binary_prob <- predict(iris_binary_fit, newdata = iris_binary_test, type = "prob")
head(iris_binary_prob)
```

We can now turn those probabilities into hard class labels using a threshold of `0.5`.

```{r iris-binary-predictions}
iris_binary_pred <- factor(
  if_else(iris_binary_prob$setosa >= 0.5, "setosa", "other"),
  levels = levels(iris_binary_test$Class)
)

confusionMatrix(iris_binary_pred, iris_binary_test$Class)
```


Because we only used two predictors, we can visualize the smooth decision surface learned by the MLP.

```{r iris-binary-boundary}
iris_binary_grid <- expand.grid(
  Petal.Length = seq(
    min(iris_binary$Petal.Length) - 0.2,
    max(iris_binary$Petal.Length) + 0.2,
    length.out = 200
  ),
  Petal.Width = seq(
    min(iris_binary$Petal.Width) - 0.2,
    max(iris_binary$Petal.Width) + 0.2,
    length.out = 200
  )
)

iris_binary_grid$prob_setosa <- predict(
  iris_binary_fit,
  newdata = iris_binary_grid,
  type = "prob"
)[, "setosa"]

ggplot() +
  geom_tile(
    data = iris_binary_grid,
    aes(x = Petal.Length, y = Petal.Width, fill = prob_setosa),
    alpha = 0.45
  ) +
  geom_contour(
    data = iris_binary_grid,
    aes(x = Petal.Length, y = Petal.Width, z = prob_setosa),
    breaks = 0.5,
    color = "white",
    linewidth = 0.7
  ) +
  geom_point(
    data = iris_binary_train,
    aes(x = Petal.Length, y = Petal.Width, color = Class),
    size = 1.8,
    alpha = 0.9
  ) +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "MLP decision surface on the binary iris task",
    subtitle = "White curve marks the 0.5 classification boundary",
    fill = "P(setosa)"
  )
```

# Multiclass Classification on Full `iris`

Now we keep the same workflow, but return to the original three-species classification problem.

The main conceptual change is that the output layer now produces one score per class instead of one probability for a yes/no outcome.

```{r iris-multiclass-data}
iris_multi <- iris

set.seed(3503)
iris_multi_idx <- createDataPartition(iris_multi$Species, p = 0.7, list = FALSE)
iris_multi_train <- iris_multi[iris_multi_idx, ]
iris_multi_test  <- iris_multi[-iris_multi_idx, ]

cat("Train:", nrow(iris_multi_train), "| Test:", nrow(iris_multi_test), "\n")
```

```{r iris-multiclass-fit}
ctrl_multi <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 1,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(3504)
iris_multi_fit <- train(
  Species ~ .,
  data = iris_multi_train,
  method = "nnet",
  metric = "Accuracy",
  preProcess = c("center", "scale"),
  trControl = ctrl_multi,
  tuneGrid = expand.grid(
    size = c(2, 4, 6),
    decay = c(0, 0.01, 0.1)
  ),
  trace = TRUE,
  maxit = 300
)

iris_multi_fit
```

```{r iris-multiclass-probabilities}
head(predict(iris_multi_fit, newdata = iris_multi_test, type = "prob"))
```

```{r iris-multiclass-eval}
iris_multi_pred <- predict(iris_multi_fit, newdata = iris_multi_test)
confusionMatrix(iris_multi_pred, iris_multi_test$Species)
```

As with the binary case, we can visualize the learned decision regions. Because this model uses 4 predictors, we draw the surface over `Petal.Length` and `Petal.Width`, while holding `Sepal.Length` and `Sepal.Width` fixed at their training medians.

```{r iris-multiclass-boundary}
iris_multi_grid <- expand.grid(
  Petal.Length = seq(
    min(iris_multi$Petal.Length) - 0.2,
    max(iris_multi$Petal.Length) + 0.2,
    length.out = 250
  ),
  Petal.Width = seq(
    min(iris_multi$Petal.Width) - 0.2,
    max(iris_multi$Petal.Width) + 0.2,
    length.out = 250
  )
)

iris_multi_grid$Sepal.Length <- median(iris_multi_train$Sepal.Length)
iris_multi_grid$Sepal.Width  <- median(iris_multi_train$Sepal.Width)

iris_multi_grid$pred_species <- predict(iris_multi_fit, newdata = iris_multi_grid)

ggplot() +
  geom_tile(
    data = iris_multi_grid,
    aes(x = Petal.Length, y = Petal.Width, fill = pred_species),
    alpha = 0.30
  ) +
  geom_point(
    data = iris_multi_train,
    aes(x = Petal.Length, y = Petal.Width, color = Species),
    size = 1.8,
    alpha = 0.9
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "MLP decision surface on the multiclass iris task",
    subtitle = "Surface in petal space, with sepal features fixed at training medians",
    fill = "Predicted class",
    color = "Observed class"
  )
```

# Regression on Sacramento House Prices

We now switch to regression. This example reuses the Sacramento case study from Topic 3.4, but the focus here is only on fitting a neural network for prediction.

The feature engineering below is the same basic preprocessing logic used in Topic 3.4: log-transform skewed variables, remove problematic rows, and add a few simple engineered features. We do not repeat the full preprocessing motivation here.

```{r sacramento-data}
data(Sacramento)

prepare_sacramento_data <- function(df) {
  df %>%
    filter(price > 0, sqft > 0, baths > 0) %>%
    drop_na(price, beds, baths, sqft, type, city, zip, latitude, longitude) %>%
    mutate(
      bedsperbath = beds / baths,
      log_price   = log(price),
      log_sqft    = log(sqft),
      latitude2   = latitude^2,
      longitude2  = longitude^2
    ) %>%
    select(
      price,
      log_price,
      beds,
      baths,
      log_sqft,
      type,
      latitude,
      longitude,
      bedsperbath,
      latitude2,
      longitude2
    )
}

set.seed(3505)
sac_idx <- createDataPartition(Sacramento$price, p = 0.75, list = FALSE)
sac_train_raw <- Sacramento[sac_idx, ]
sac_test_raw  <- Sacramento[-sac_idx, ]

sac_train <- prepare_sacramento_data(sac_train_raw)
sac_test  <- prepare_sacramento_data(sac_test_raw)

cat("Train after cleaning:", nrow(sac_train), "| Test after cleaning:", nrow(sac_test), "\n")
```

And we build the desired predictor formula

```{r sacramento-formula}
sac_formula <- log_price ~ beds + baths + log_sqft + type +
  latitude + longitude + bedsperbath + latitude2 + longitude2
```


For regression we keep the same `size` and `decay` tuning idea, but tell `nnet` to use a linear output with `linout = TRUE`.

```{r sacramento-nn-fit}
ctrl_reg <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 1
)

set.seed(3506)
sac_nn_fit <- train(
  sac_formula,
  data = sac_train,
  method = "nnet",
  metric = "RMSE",
  preProcess = c("center", "scale"),
  trControl = ctrl_reg,
  tuneGrid = expand.grid(
    size = c(3, 5, 7),
    decay = c(0, 0.01, 0.1)
  ),
  trace = TRUE,
  linout = TRUE,
  maxit = 500,
  MaxNWts = 2000
)

sac_nn_fit
```

```{r sacramento-nn-eval}
sac_nn_pred_log <- predict(sac_nn_fit, newdata = sac_test)

metrics_table <-
  as_tibble_row(postResample(pred = sac_nn_pred_log, obs = sac_test$log_price)) %>%
    mutate(Model = "Neural network", .before = 1)

metrics_table
```

The model is trained on `log_price`, but for interpretation we plot predictions back on the original dollar scale.

```{r sacramento-predicted-vs-observed}
sac_results <- tibble(
  observed_log = sac_test$log_price,
  predicted_log = sac_nn_pred_log,
  observed_price = exp(observed_log),
  predicted_price = exp(predicted_log)
)

ggplot(sac_results, aes(x = observed_price, y = predicted_price)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "firebrick") +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(
    title = "Sacramento test set: observed vs predicted prices",
    subtitle = "Predictions come from an MLP trained on log(price)",
    x = "Observed price",
    y = "Predicted price"
  )
```


# Final Remarks

We have seen a very short and simple way of fitting small NNs with packages that do not allow tuning all the internal possibilities that a neural network has. If any one is in that, there are several tutorials on the [torch for R webpage](https://torch.mlverse.org/) 
