---
title: "0 - Breaking the ice_SOLVED"
course: Text Mining - MUCSS 25/26
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
instructor: Carmen Torrijos
editor_options: 
  chunk_output_type: inline
---

# Breaking the ice

This is a little exercise to start learning how to deal with text.

## 1. Choose a short text of your liking (5-6 lines)

It can be a Lord Byron poem, a Bad Bunny song, a Trump speech or just a short joke in English. In my case, it's a verse in Ed Sheeran's song *Give me love*.

## 2. Save it into a variable

Break it into lines and put it in quotes. **In code, text must always be in quotes**.

```{r}
my_text <- c("Give me love like never before",
          "'Cause lately I've been craving more",
          "It's been a while but I still feel the same,",
          "Maybe I should let you go")
```

Print your variable to see everything is OK.

```{r}
my_text
```

## 3. Let's make a tibble

To make it tidy, let's convert this into a tibble. **A tibble is a modern class of data frame in R**, available in the `dplyr` and `tibble` packages, that has a beautiful print method and keeps things organized.

```{r}
library(dplyr)
#tibble() has two parameters: the lines we need (line) and the variable containing the text (text)
my_text_df <- tibble(line = 1:4, text = my_text)

my_text_df
```

Here we have **one sentence per row**. We still can not analyze too much in this format.

## 4. Let's break it into words

A **token** is a meaningful unit of text, most often a [word]{.underline} but not only, that we are interested in using for further analysis.

**Tokenization** is the process of splitting text into tokens. Use the function `unnest_tokens()` to make our tibble contain a new column with tokens.

```{r}
# install.packages("tidytext")
```

```{r}
#call the library tidytext
library(tidytext)

my_text_df %>%
  #The two basic arguments to unnest_tokens are the new column name (word) and the column tokens come from (text)
  unnest_tokens(word, text)
```

Note that default tokenization for this function is:

-   punctuation has been stripped

-   it converts the text to lowercase

-   the sentence each word comes from is not shown

-   and only single words (like *stop*), not multiple words (like *stop_for*)

## 5. Let's use some options

Use the [function documentation](https://www.rdocumentation.org/packages/tidytext/versions/0.4.1/topics/unnest_tokens) to find arguments that solve the following problems:

**5.1 Source line**

We want to create the tokenization tibble knowing which line each word tokenized came from. Look for the correct argument and modify the dataframe.

```{r}
my_text_df %>%
  #find the argument for keeping the line the word came from
  unnest_tokens(word, text, drop=FALSE)
```

**5.2** **Punctuation**

Maybe we want to have our punctuation symbols treated as tokens too. Look for the correct argument and modify the dataframe again.

**Note**: If -by any chance- your text has no punctuation symbols, please add some manually to see the difference.

```{r}
my_text_df %>%
  #drop argument is for keeping the line the word came from
  unnest_tokens(word, text, drop = FALSE, strip_punct = FALSE)
```

**5.3** **Case**

And finally, we want to respect the case.

**Note**: If -by any chance- your text has no upper case letters, please add them manually to see the difference.

```{r}
my_text_df %>%
  #drop argument is for keeping the line the word came from
  unnest_tokens(word, text, drop = FALSE, strip_punct = FALSE, to_lower = FALSE)
```

Congrats! You have learned the basics for a tokenization process over one text.
