---
title: "Logistic Regression"
format: html
editor: visual
---

## Load data from mlbench package

```{r}
data("PimaIndiansDiabetes2", package = "mlbench")
library(caret)
library(tidyverse)

# we remove rows with NA values
df <- na.omit(PimaIndiansDiabetes2)
```

Variables info:

-   pregnant: Number of times pregnant
-   glucose: Plasma glucose concentration (glucose tolerance test)
-   pressure: Diastolic blood pressure (mm Hg)
-   triceps: Triceps skin fold thickness (mm)
-   insulin: 2-Hour serum insulin (mu U/ml)
-   mass: Body mass index
-   pedigree: Diabetes pedigree function (metric of genetic predisposition)
-   age: Age (years)
-   diabetes: detected diabetes (yes or no) --\> target variable

```{r}
summary(PimaIndiansDiabetes2)
```

# Perform Exploratory Data Analysis

You can use DataExplorer hereas we've done before!

Train logistic regression

```{r}
#split train and test datasets
training.samples <- df$diabetes |> 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- df[training.samples, ]
test.data <- df[-training.samples, ]
```

```{r}
# Fit the model
model <- glm(diabetes ~., data = train.data, family = binomial)
summary(model)
```

```{r}
round(exp(coef(model)),2)
```

The exponential of the slope coefficient (exp(B)) tells us the factor by which odds increase if the independent variable increases by one unit.

-   One unit increase in glucose levels increase the odds of diabetes by 1.04, this is a 4% increase in the odds of being diabetic (1.04 \* odds)

*Provide the interpretation for the following*: - pregnant - pressure - pedigree

Make predictions and check general accuracy

```{r}
# Predictions
probabilities <- model |> predict(test.data, type = "response")

# select the classification threshold
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
```

Model graph

```{r}
train.data |>
  mutate(prob = ifelse(diabetes == "pos", 1, 0)) |>
  ggplot(aes(glucose, prob)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(
    title = "Logistic Regression Model", 
    x = "Plasma Glucose Concentration",
    y = "Probability of being diabete-pos"
    )
```

# Model evaluation

Confusion Matrix & derived metrics

```{r}
confusionMatrix(as.factor(test.data$diabetes), as.factor(predicted.classes))
```

ROC & AUC

```{r}
library(pROC)
#creates an object with all sorts of diagnostics 
roc <- roc(as.factor(test.data$diabetes), probabilities) 

pROC::ggroc(roc)

# manual option 
sens <- roc$sensitivities #include sensitivities in test data
spec <- roc$specificities

ggplot(data.frame(sens,spec), aes(x=1-spec, y=sens)) + geom_line()
```

Area Under the Curve

```{r}
auc(test.data$diabetes, probabilities)
```

# Exercise

Evaluate the model using the train set. How do performance metrics change?

```{r}
# Predictions
probabilities <- model %>% predict(train.data, type = "response")

# select the classification threshold
predicted.classes <- ifelse(probabilities > 0.5, "pos", "neg")
```

```{r}
confusionMatrix(as.factor(train.data$diabetes), as.factor(predicted.classes), positive = "pos")
```

```{r}
#creates an object with all sorts of diagnostics 
roc <- roc(as.factor(train.data$diabetes), probabilities) 

pROC::ggroc(roc)
```

```{r}
auc(train.data$diabetes, probabilities)

```
