---
title: "Poisson Regression"
format: html
editor: visual
---

Libraries

```{r}
library(ggplot2)
library(dplyr)
```

```{r}
# Example: Website daily visitor counts
data <- data.frame(
  visitors = c(42, 48, 39, 52, 44, 58, 51, 47, 41, 49, 40, 46, 43, 54, 50),
  day_of_week = factor(c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun",
                         "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun", "Mon")),
  marketing_spend = c(200, 150, 180, 250, 220, 400, 350, 300, 160, 190,
                      140, 210, 380, 420, 170),
  temperature = c(22, 25, 19, 24, 26, 28, 30, 27, 21, 23, 18, 25, 29, 31, 20)
)
```

```{r}
# Check mean vs variance (should be similar for Poisson)
mean(data$visitors)
var(data$visitors)

# variance to  mean ratio 
var(data$visitors) / mean(data$visitors) 
# Should be close to 1 -- Ratios between 0.5 and 1.5 are generally acceptable 
```

The ratio of variance to mean is low means we have a bit of under-dispersion.

```{r}
# Visualize the distribution
ggplot(data, aes(x = visitors)) +
  geom_histogram(bins = 8, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Daily Visitors",
       x = "Number of Visitors", y = "Frequency")
```

Since it is just 15 observations, it does not look like a poisson.

```{r}
# Basic Poisson regression
model <- glm(visitors ~ day_of_week + marketing_spend + temperature,
             family = poisson, data = data)

# View key model results
summary(model)
```

Similar to lms.

```{r}
# Calculate exponentiated coefficients - Incidence Rate Ratios (IRRs)
exp(coefficients(model))

```

Taking the exponential of the coefficients in order to interpret them.

```{r}
# Create new data for a prediction
new_data <- data.frame(
  day_of_week = factor("Fri", levels = levels(data$day_of_week)),
  marketing_spend = 300,
  temperature = 25
)

```

Create a prediction for a specific day. Predict the number of visitors given the three "regressors" or variables.

```{r}

# Predict expected counts
predicted_counts <- predict(model, newdata = new_data, type = "response")
print(paste("Expected visitors:", round(predicted_counts, 1)))

```

Run "predict" to create a predicted value. Note it is not an integer.

# Model Evaluation

We check for overdispersion with the deviance ratio

```{r}
# Calculate the dispersion statistic
residual_deviance <- model$deviance
df_residual <- model$df.residual
dispersion <- residual_deviance / df_residual

dispersion

if (dispersion > 1.5) {
  print("Possible overdispersion detected")
  print("Consider quasi-Poisson or negative binomial models")
}
```

It is very close to one, so we don't have a problem of overdispersion.

We can also run a test Ho: There is no overdispersion

```{r}
# Overdispersion test
library(AER)
dispersiontest(model)
```

Does not seem there is overdispersion (high p-value)

```{r}
# Calculate Pearson residuals
fitted_values <- fitted(model)
pearson_residuals <- residuals(model, type = "pearson")

# Plot residuals vs fitted values
plot(fitted_values, pearson_residuals,
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```

If these are more or less scattered randomly across the line, we do not have a problem of incorrect fitting.
