---
title: "Stargazer example"
subtitle: "Survey Research Methodology II"
author:
  - name: "Marga Torre"
    email: "margarita.torre@uc3m.es"
format:
  html:
    toc: true
    toc-location: left
    theme: 
    - united
    self-contained: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
lang: en
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T, message=FALSE, warning=FALSE, knitr.purl.inline = TRUE )
```

## Example: Determinants of Low Self-Esteem

In this example, we estimate different logistic regression models to explain **low self-esteem**.

The dependent variable is binary:

-   `1 = low self-esteem`
-   `0 = otherwise`

We progressively add groups of predictors to compare model specifications.

# Generate synthetic survey data

```{r}
library(tidyverse)
library(stargazer)

set.seed(123)

n <- 1200

df <- tibble(
  age = round(rnorm(n, mean = 35, sd = 12)),
  female = rbinom(n, 1, 0.55),
  education = sample(1:5, n, replace = TRUE,
                     prob = c(0.15, 0.25, 0.30, 0.20, 0.10)),
  unemployed = rbinom(n, 1, 0.10),
  income = round(rlnorm(n, meanlog = 10.2, sdlog = 0.5)),
  social_media_hours = round(pmax(rnorm(n, 2.5, 1.5), 0), 1),
  bullying_experience = rbinom(n, 1, 0.20),
  social_support = round(pmin(pmax(rnorm(n, 6, 2), 0), 10), 0)
) %>%
  mutate(
    log_income = log(income),
    linpred = -2.0 +
      0.30 * female +
      0.60 * unemployed -
      0.20 * education -
      0.15 * log_income +
      0.25 * social_media_hours +
      0.90 * bullying_experience -
      0.35 * social_support +
      rnorm(n, 0, 0.5),
    low_self_esteem = rbinom(n, 1, plogis(linpred))
  )
```

Check the df

```{r}
head(df)
```

### Model Specifications

We estimate three logistic regression models:

-   **Model 1**: Demographic variables\
-   **Model 2**: Adds socioeconomic characteristics\
-   **Model 3**: Adds psychosocial factors

```{r}
# Model 1: Demographics
m1 <- glm(low_self_esteem ~ age + female,
          data = df, family = binomial)
summary(m1)
# Model 2: Add socioeconomic variables
m2 <- glm(low_self_esteem ~ age + female + education + unemployed + log_income,
          data = df, family = binomial)
summary(m2)
# Model 3: Add psychosocial factors
m3 <- glm(low_self_esteem ~ age + female + education + unemployed + log_income +
            social_media_hours + bullying_experience + social_support,
          data = df, family = binomial)
summary(m3)
```

### Comparing Models with Stargazer

The table below compares the three specifications. Coefficients are reported in **log-odds**, with standard errors in parentheses.

```{r}
stargazer(
  m1, m2, m3,
  type = "text",  
  title = "Determinants of Low Self-Esteem (Logit Models)",
  dep.var.labels = "Low Self-Esteem (1 = yes)",
  column.labels = c("Model 1", "Model 2", "Model 3"),
  covariate.labels = c(
    "Age",
    "Female",
    "Education (1-5)",
    "Unemployed",
    "Log Income",
    "Social Media Hours",
    "Experienced Bullying",
    "Social Support (0-10)"
  ),
  digits = 3,
  star.cutoffs = c(0.05, 0.01, 0.001),
  notes = "Logit coefficients (log-odds). Standard errors in parentheses."
)
```

In the table:

-   Each column corresponds to a different model specification
-   Each row shows the estimated effect of a predictor
-   Standard errors are reported in parentheses below the coefficients
-   Stars indicate statistical significance (\* p \< 0.05; \*\* p \< 0.01; \*\*\* p \< 0.001)

Stargazer offers many additional customization options that can be explored in the official documentation. For example, you can display odds ratios instead of log-odds (`apply.coef = exp`), omit specific statistics (`omit.stat = c("f", "ser")`), rename variables (`covariate.labels = ...`), change significance thresholds (`star.cutoffs = ...`), or export tables directly to HTML or LaTeX (`type = "html"` or `type = "latex"`).
